### 2025-05-14 articles/arxiv/paper.tex
upload draft of paper from overleaf

\documentclass{article}

\usepackage{authblk, array, float, graphicx, hyperref, natbib, subcaption, amsmath}
\usepackage[margin=1in]{geometry}

\bibliographystyle{../apa-good}
\linespread{1.2}

\title{A comps-based approach for interpreting tree-based predictions with an application to NFL quarterback draft prospect projection}
\author[1]{Elisabeth Millington}
\author[2]{Scott Powers}
\affil[1]{Department of Kinesiology, Rice University}
\affil[2]{Department of Sport Management, Rice University}

\begin{document}

\maketitle

\begin{abstract}
  The abstract goes here.
\end{abstract}

\section{Introduction}

Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to control for overfitting. Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks). One of the primary strengths of random forest models are their robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.

One drawback of random forest models is that the reasons for an individual's prediction can be difficult to interpret. Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it allows the user to trust the prediction \citep{ribeiro_why_2016}.  We introduce a method for interpreting random forest predictions that connects directly to the concept of player comparisons, which is a commonly used framework in traditional scouting. % improve this

Quarterbacks are the backbone of offensive schemes for every National Football League (NFL) team and are arguably the most important position \citep{hughes_positional_2015}. Due to the nature of the position, trading for/signing a solid quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success. Since quarterbacks are highly sought after, talented college quarterbacks with perceived NFL starting potential are typically taken in the first or second round of the draft. In the NFL, this can have significant consequences because there is a set salary that increases for each higher draft pick. This is one of many reasons front offices must make informed decisions when drafting a quarterback.

One method that teams use when evaluating college players is comparison to successful NFL players. Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers. These comparisons are not exact, but they are meant to inform fans and scouts of player potential in a more digestible way and are known as ``NFL comps'' \citep{jones_nfl_2025}.

In this paper, we aim to use a random forest model to generate similarity scores between quarterbacks in college football and professional players, with the ultimate goal of adding interpretability to the random forest model's predictions. Using ESPN's Total Quarterback Rating as the response variable and college statistics as the predictor variables, we can use the relationship between random forests and \textit{k}-nearest neighbor algorithm \citep{lin_random_2006} to create a formula for similarity.

While traditional scouting relies on subjective observations, these similarity scores aim to reduce the uncertainty in draft evaluation. Using quantifiable data allows for a more objective measure of identifying players who may follow a similar career trajectory. This approach offers a data-driven perspective on evaluating draft prospects, which adds to the subjective data provided by scouts.

In this paper, we propose a methodology for calculating player similarity scores using a random forest model. We then apply this methodology to the quarterback position, focusing on how performance metrics from college football can be used to assess a player's potential at the professional level. By examining these scores, we provide a tool to improve the accuracy of draft projections and support more informed decision-making processes.

\subsection{Related Work}
% Section 1.1 could use a little more narrative flow. You jump right in describing an article in the first sentence, but it would be helpful to provide more context. Is that article about sports? I think this section could be a little better organized into paragraphs with each paragraph representing one topic (e.g. interpreting machine learning algorithms, predicting player success in the draft, the concept of comps, etc.), with a sentence at the top of each paragraph explaining/introducing the topic. It would also be good to end this section with a paragraph explaining what your work accomplishes that hasn't been done before.

Previous research has been done to attempt to introduce interpretability to random forest predictions. \citet{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner. \citet{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models. In a sports context, \citet{metulini_measuring_2023} use generalized Shapely values to interpret individual player contributions based on a win probability model in the context of professional basketball.

Draft models are fairly common in professional sports, as evaluating prospect talent can be key to success. One example of a draft model that uses QBR as the response variable is \citet{craig_predicting_2021}. They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season's metrics, scout grades, and player height. Their predictions are easy to interpret directly from the logistic regression model. Another study uses their own metric of net points combined with games played to define NFL success. \citet{wolfson_quarterback_2011} use a combination of NFL Combine and college statistics to create a draft model. They use this model to analyze how effectively NFL teams leverage data available when drafting a quarterback. They use a logistic regression model, which results in predictions that are easily interpretable. \citet{berger_jumping_2021} uses principal component analysis (PCA) combined with a high order regression model to evaluate NBA prospects using their college statistics and Combine data. The reasoning behind the PCA predictions are unclear, resulting in a relatively uninterpretable draft model. \citet{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics. They use a combination of large language models and ensemble machine learning, which are both opaque in the way that they generate predictions, leading to uninterpretable predictions/reports.

There are several articles from various sports reporters that discuss player comps from analysts. Pro Football Focus published their comps for their top 20 prospects this year, however the only quarterback prospect on the list was Cam Ward, whom they compared to Ben Roethlisberger \citep{sikkema_2025_2025}. The Giants also published an article with various analyst opinions on top prospect comps. This article included three quarterbacks: Cam Ward, Shedeur Sanders, and Jalen Milroe \citep{noauthor_draft_2025}. A final example is an article from CBS Sports that describes current prospect comps to past NFL busts \citep{trapasso_nfl_2025}.

While the articles discussed show the previous work that has been done to interpret machine learning models, evaluate prospects, and make player comparisons, none of them have combined these three concepts. This paper introduces interpretability to random forest predictions in the context of sports, specifically American Football. In doing so, we also introduce a method of calculating quantifiable player comps between the training players in the dataset and the prospects.

\section{Data}

There were two sources for our data. The NFL data for ESPN's Total Quarterback Rating (QBR) came from \href{https://pro-football-reference.com}{pro-football-reference.com}. For the corresponding college statistics, we used data from \href{https://sports-reference.com}{sports-reference.com}. The college statistics only include schools that are a part of the NCAA Division 1 Football Bowl Subdivision (FBS). We were able to obtain college statistics dating back to 1984. However, we were constrained by the ESPN QBR metric, which has calculations starting at the 2006 season.

The data from Sports Reference contained player passing statistics for each season. It contained categories for player, team, conference, games, completions, attempts, yards, touchdowns, touchdown percentage, interceptions, interception percentage, yards per attempt, adjusted yards per attempt, yards per completion, yards per game, passing efficiency rating, awards, and season. In addition to the passing statistics, we included player rushing statistics for each season from Sports Reference.

We obtained NFL season passing statistics from Pro Football Reference for the 2006-2024 seasons. The resulting data frame contained columns for 28 different passing metrics. However, we only used QBR and overall pick from this model.

\subsection{ESPN's Total Quarterback Rating}

The NFL response variable we used in this paper is ESPN's Total Quarterback Rating (QBR). It is an EPA-based metric that is adjusted for home-field advantage, defensive strength, and ``garbage time'' and then scaled from 0-100 \citep{noauthor_how_2016}. QBR provides a more holistic evaluation of quarterbacks in the NFL while also considering other factors affecting performance. Additionally, a 0-100 scale is important in this context because it offers an interpretable standard frame of reference. It also allows us to assign a meaningful value to quarterbacks who never played in the NFL, compared to raw EPA (where the lack of such scale results in negative values for some players). For example, it would not make much sense for a quarterback who receives playing time in the NFL to have a lower value than a quarterback who never made it to the NFL, but this would happen if using raw EPA as the response variable. Using QBR as the response variable addresses this issue while incorporating other important aspects of game-time performance.

\section{Methods}

\subsection{Data Preparation}

\subsubsection{Data Cleaning}

In order to clean the data frame containing college statistics, we started by joining the rushing dataframe with the passing dataframe by player and season. We transformed the data frame to contain only one player per row by taking the average of their games played, completions, attempts, passing yards, passing passing touchdowns, interceptions, rushing attempts, rushing yards, and rushing touchdowns. In addition to career statistics, we added columns for the career games played, career seasons played, career yards per attempt, final season's yards per attempt, final season, final season's games played, final season's passer rating, final season's college, and the final season's conference.  Next, we created a binary variable for winning the Heisman Award. We also created a variable for how each player finished in the Heisman voting in their final college season and how many years the player was an All-American selection. Finally, we filtered the dataset to remove players who had less than 5 average attempts per season and players with less than less than 6 career games and whose last season was before 2024 (as not to include players still in college).

In the NFL dataset, we calculated the average QBR per season played for each player. The resulting data frame contained one row per player with columns for player name and average QBR. We combined this data frame with the college data frame and replaced any missing QBR values (due to never having played in the NFL) with zero. We also filtered the data to remove any players whose final year in college was before 2000 since our NFL data only dates back to the 2006 season. Our final dataset included statistics for 2,115 quarterbacks.

\subsection{Random Forest Model}

Random forests are a machine learning method typically for classification or regression. For regression tasks, \textit{n} decision trees are created during training. Each tree produces a prediction of the response variable, and the final prediction is the average of the individual trees' predictions. The use of multiple random decision trees reduces overfitting by introducing more bias, which reduces variance.

We trained the random forest model using the R package ranger, which is a fast implementation of the random forest algorithm \citep{wright_ranger_2015}. The response variable was average QBR per season (zero for players who never appeared in the NFL). The college statistics used as predictor variables were career games played, average completions per season, average attempts per season, average yards per season, average touchdowns per season, average interceptions per season, average rushing attempts per season, average rushing yards per season, average rushing touchdowns per season, career yards per attempt, career All-American designations, final season yards per attempt, final season games played, final season passer efficiency rating, final season Heisman award voting, whether the player won the Heisman award, and final season NCAA conference. We implemented a random 70\%/30\% training/testing data split, and the random forest contained 500 decision trees.

To determine the best hyperparameters, we used the tuneRanger function from the tuneRanger package \citep{probst_tuneranger_2018} with 100 iterations and RMSE as the measure to optimize. We found the best parameters to be \texttt{min.node.size} = 41 and \texttt{mtry} = 4.

\subsection{Similarity Score}

\subsubsection{\textit{k}-Nearest Neighbors}

\citet{lin_random_2006} previously established that there is a relationship between \textit{k}-nearest neighbors and random forest. Given that the training data is $\{(x_i,y_i),i=1, ...,n\}$, a prediction for $x_0$ from any given tree in a random forest model is $$\hat{y}_m(x_0)=\sum_{i=1}^{n}W_{im}y_i$$
where:
\[W_{im} = 
\begin{cases}
\frac{1}{k_m}, & \text{if } x_i \text{ is in the same terminal node as } x_0 \text{ in tree } m, \\
0, & \text{otherwise}.
\end{cases} 
\]
and $k_m$ denotes the number of training points in that terminal node.

This setup defines a local neighborhood around $x_0$ for tree $m$.

The final random forest prediction results from averaging the individual tree predictions $$\hat{y}(x_0)=\sum_{i=1}^M\bar{W_i}y_i$$
where the average weight, $\bar{W_i}$, is defined as $$\frac{1}{M}\sum_{m=1}^MW_{im},$$ where $M$ is the total number of trees in the forest.

This formulation shows that a random forest can be interpreted as a form of \textit{k}-NN, where the forest itself is a weighted neighborhood scheme and neighborhoods are defined by tree structure rather than distance. The training points that share a leaf node with $x_0$ in tree $m$ are considered its neighbors. The final prediction results from averaging the results across all trees, where training points that show up as neighbors more frequently receive higher weight.

\subsubsection{Similarity Formula}
Using the random forest model's relationship to \textit{k}-NN, we can derive a formula for similarity scores between testing player \textit{i} and training player \textit{j}. We derived the formula for the similarity score between player $i$ and player $j$ to be
$$S(i,j) = \frac{1}{M}\sum_{k=1}^{M}{\frac{N_{j,T_{i,k}}}{|T_{i,k}|}}$$
where:
\begin{itemize}
    \item $|T_{i,k}|$ is the total number of players in player \textit{i}'s leaf nodes of tree \textit{k} 
    \item $N_{j,T_{i,k}}$ is the number of times player \textit{j} appears in player \textit{i}'s leaf nodes in tree \textit{k}
\end{itemize}

We can use these similarity scores to recreate the predictions for player $i$ from the random forest by taking the weighted average of the training players' observed performance using the similarity scores as the weight.  Note that for each tree $k$, the weight $W_{jk}$ is $\frac{1}{|T_{i,k}|}$ if $j \in T_{i,k}$ and $0$ otherwise. Therefore, player $j$'s contribution across all trees can be defined as

$$\bar{W}_j = \frac{1}{M} \sum_{k=1}^M \frac{N_{j,T_{i,k}}}{|T_{i,k}|} = S(i,j),$$

and substituting back into the prediction formula, we obtain:
\[
\hat{y}_i = \sum_{j=1}^n \bar{W}_j y_j = \sum_{j=1}^n S(i,j) y_j.
\]
The similarity score $S(i,j)$ serves as the prediction weight for training player $j$ when predicting the output for test player $i$.

\section{Results}

\subsection{Model Performance}

The model's test root mean squared error (RMSE) is 9.88, and 33.87\% of the model's variance is explained. A plot of the variable importance is below.

\begin{figure}[H]
    \centering
%    \includegraphics[width=0.75\linewidth]{model plots/variable_importance.png}
    \caption{\textit{A histogram showing the importance of each predictor variable in the random forest model predicting career average QBR.}}
    \label{fig:importance}
\end{figure}

\begin{figure}[H]
    \centering
%    \includegraphics[width=0.75\linewidth]{model plots/3d_plot.png}
    \caption{\textit{This plot shows the predictions on the testing set as a function of college touchdowns per season and final season passer rating. The color of the point represents the QBR prediction value on a continuous scale, while the x-axis represents college yards per season, and the y-axis represents the final season passer rating.}}
\end{figure}

Figure \ref{fig:predicted-vs-actuals} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft. Cam Ward is predicted to have the highest QBR, which matches the common consensus that he is the best quarterback in this draft class and the fact he was taken first overall in the draft. Dillon Gabriel and Shedeur Sanders were predicted second and third respectively. While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he has had a very consistent college career and finished the highest out of any quarterback in Heisman voting this year.

\begin{figure}[H]
    \centering
%    \includegraphics[width=0.75\linewidth]{model plots/predicted_vs_actuals.png}
    \caption{\textit{Actual QBR plotted against the random forest generated QBR prediction for the testing dataset.}}
    \label{fig:predicted-vs-actuals}
\end{figure}

\input{../../tables/top_ten.tex}

\subsection{2025 NFL Draft Comps}

Table 2 below shows the result of calculating similarity scores for the four chosen quarterback prospects. These scores are essentially the percentage of the prospect's leaf nodes that are shared with that training player. We can use these to interpret the predictions from Table 1 as the weights each of the training players has on the prospect's QBR prediction. We can see that for Cam Ward, Shedeur Sanders, and Dillon Gabriel, the vast majority of the training players with the top ten similarity scores are relatively successful starting (or formerly starting) quarterbacks. Intuitively, since they had the three highest predictions for this draft class, it would make sense that their predictions are based on players who are/were successful in the NFL (i.e. had a high career average QBR). For Jaxson Dart, we can see that only four of the ten players who have the highest similarity scores were ever starters in the NFL. This means that the other six players likely have a QBR value that is close to zero, which explains why his prediction is significantly lower than the other three prospects.

These four quarterbacks were selected to illustrate the model’s interpretability and its alignment with real-world outcomes. We used Shedeur Sanders and Cam Ward as two examples because they were both initially expected to go in the first round of this year's draft. Shedeur Sanders' draft position did unexpectedly slide, and he ended up being taken in the fifth round in the draft. However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance \citep{mckenna_what_nodate}. We chose Jaxson Dart because his draft stock continuously increased after the NFL Scouting Combine and his Pro Day, and he was one of the two quarterbacks taken in the first round of the NFL draft. Finally, we chose Dillon Gabriel because he finished the highest of all quarterbacks in Heisman voting this year (3rd overall). We used a model containing the full dataset (not just the training data) to calculate the similarity scores for each of these four players using the above-described methodology. It is worth noting that both Cam Ward and Shedeur Sanders started their college careers at teams that are not in the NCAA D1 FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State. This means that their college statistics used in the model do not include their statistics from these schools. The table below serves as a tool for interpreting the above random forest predictions for these draft prospects. The percentages offer a quantifiable measure of similarity based on college statistics that underlie the random forest prediction output, allowing us to interpret the predictions better. We can see that Cam Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class. Dillon Gabriel has a slightly higher QBR prediction than Shedeur Sanders, even though Shedeur was considered a better prospect. However, Gabriel finished higher than Shedeur in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft. Jaxson Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration. He also performed well in his interviews with teams, which is not a quantifiable measure.

\begin{table}[H]
\resizebox{\textwidth}{!}{
  \input{../../tables/side_by_side_similarity.tex}}
  \caption{\textit{A table of the top similarity percentages for each of the four selected prospects. The similarity percentage is the percentage of nodes that the training quarterback and prospect ended in the same node based on the predictor variables.}}
\end{table}

Figure \ref{fig:prospect-plots} shows the weights of the QBRs used to calculate the predictions, with a green line representing the value of the prediction. There is also a graph that shows the similarity scores vs. QBR value. These plots were produced for each of the four 2025 prospects. The scatterplots show the training player similarity score against their QBR for each of the prospects. We can see that there is a large concentration of training players with a value of 0 for their QBR value. This results from the training dataset containing 2,114 players, with only 300 of those players recording NFL stats (meaning the rest have a value of 0 for the mean QBR). We can see that besides the high concentration of training players at zero, Cam Ward has the highest concentration of players around 55-65, which is why his prediction is the highest. On the opposite end of the spectrum, Jaxson Dart has the highest concentration of training players at around zero and the lowest concentration at relatively high QBR values. This explains why his prediction was by far the lowest of the four prospects we looked at.

\begin{figure}[H]
    \centering
%    \includegraphics[width=0.24\linewidth]{prospect plots/cw_plot1.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/ss_plot1.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/dg_plot1.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/jd_plot1.png} \\
%    \includegraphics[width=0.24\linewidth]{prospect plots/cw_plot2.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/ss_plot2.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/dg_plot2.png}
%    \includegraphics[width=0.24\linewidth]{prospect plots/jd_plot2.png}
    \caption{\textit{Top: Histogram of training players’ QBR values weighted by the proportion of shared nodes with the prospect (similarity). The green vertical line represents prospect’s random forest predicted QBR based on the similarity-weighted average. The red density curve provides a smoothed visualization of the distribution.
    Bottom: Scatter plot showing similarity scores for training players plotted against their actual QBR values. Higher points indicate greater similarity to the prospect, highlighting which historical players he shared the nodes with in the random forest model.}}
    \label{fig:prospect-plots}
\end{figure}

\section{Discussion}

We have demonstrated a methodology that produces similarity scores between draft prospects and current/past NFL players based on college statistics by using the nearest-neighbors properties of the random forest algorithm. Comparisons between draft prospects and current players are important because they allow the analytical models to be presented in a more digestible way to fans and front offices. They allow front offices to make more informed decisions on draft day. Additionally, this method provides a way to interpret and understand the random forest algorithm.

This methodology is not specific to quarterbacks or even professional football. There are further applications for other position groups, as well as other sports. Similarity scores can be used when scouting running backs, wide receivers, defensive backs, linebackers, or virtually any position with well documented college statistics. Beyond football, this methodology can be adapted for basketball, baseball, or soccer and can even be used when considering bringing up players from the minor leagues.

This model's performance means that close to one third of the variance in QBR is explained by these variables, which highlights some limitations of statistical forecasting when scouting draft prospects. The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance. However, these two things also highlight the importance of non-quantifiable factors – such as attitude, coachability, off-the-field issues, adaptability, and leadership.

One key strength of this approach to comps is its objectivity. It enables teams to leverage historical data to make comps, without relying solely on scouting narratives or subjective analysis. While this approach should not replace traditional scouting, it can complement the existing strategies and provide additional information to front offices.

\bibliography{../random-forest-comps}

\end{document}

### 2025-05-14 articles/arxiv/paper.tex
fine-tune introduction section

@@
 
-\title{A comps-based approach for interpreting tree-based predictions with an application to NFL quarterback draft prospect projection}
+\title{A comps-based approach for interpreting tree-based predictions}
 \author[1]{Elisabeth Millington}
@@
 
-Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to control for overfitting.
+Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to mitigate overfitting.
 Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks).
-One of the primary strengths of random forest models are their robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.
+Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.
 
-One drawback of random forest models is that the reasons for an individual's prediction can be difficult to interpret.
-Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it allows the user to trust the prediction \citep{ribeiro_why_2016}.
-We introduce a method for interpreting random forest predictions that connects directly to the concept of player comparisons, which is a commonly used framework in traditional scouting.
-% improve this
+Like other so-called ``black box'' machine learning algorithms, one drawback of the random forest is that the reasons for an individual's prediction can be difficult to interpret.
+Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction \citep{ribeiro_why_2016}.
+This is especially important in sports contexts because executives need to consider non-quantifiable information in their decision-making.
+The better a user understands how a model arrives at a prediction, the better they can determine how much it is appropriate to adjust the prediction based on subjective information from outside of the model.
 
-Quarterbacks are the backbone of offensive schemes for every National Football League (NFL) team and are arguably the most important position \citep{hughes_positional_2015}.
-Due to the nature of the position, trading for/signing a solid quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
+Perhaps one of the most important decisions that sports executives make is selecting quarterbacks in the National Football League (NFL) draft.
+Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position \citep{hughes_positional_2015}.
+Due to the nature of the position, trading for or signing a solid quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
 Since quarterbacks are highly sought after, talented college quarterbacks with perceived NFL starting potential are typically taken in the first or second round of the draft.
-In the NFL, this can have significant consequences because there is a set salary that increases for each higher draft pick.
-This is one of many reasons front offices must make informed decisions when drafting a quarterback.
+In the NFL, this can have significant consequences because the rookie salary scale means large financial commitments to players taken at the top of the draft.
+In 2025, the slotted contract for No.
+1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus \citep{badenhausen_nfl_2025}.
 
-One method that teams use when evaluating college players is comparison to successful NFL players.
+In traditional scouting for sports, draft prospect discussions and predictions are often based on comparing a current prospect with similar draft prospects in the past.
+Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success \citep{trapasso_nfl_2025}.
 Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers.
-These comparisons are not exact, but they are meant to inform fans and scouts of player potential in a more digestible way and are known as ``NFL comps'' \citep{jones_nfl_2025}.
+These historical reference players are known colloquially as ``comps'' \citep{jones_nfl_2025}, short for comparables.
+The simplest example of this line of reasoning is, ``This prospect reminds me most of Player X, and Player X had a successful career, so I think this player will have a successful career,'' effectively a human implementation of the $k$-nearest neighbor ($k$-NN) algorithm, with $k = 1$.
 
-In this paper, we aim to use a random forest model to generate similarity scores between quarterbacks in college football and professional players, with the ultimate goal of adding interpretability to the random forest model's predictions.
-Using ESPN's Total Quarterback Rating as the response variable and college statistics as the predictor variables, we can use the relationship between random forests and \textit{k}-nearest neighbor algorithm \citep{lin_random_2006} to create a formula for similarity.
+Random forests can be thought of as an adaptive $k$-NN algorithm \citep{lin_random_2006}, with the nearness of neighbors adaptively determined to minimize prediction error.
+Our work is moivated by this connection between random forests and the traditional lens of player comps.
+In this paper, we develop methods and software for extracting similarity scores from fitted random forest models in such a way that the prediction for a new data point is the weighted average of the outcomes in the training set, weighted according to these similarity scores.
 
-While traditional scouting relies on subjective observations, these similarity scores aim to reduce the uncertainty in draft evaluation.
-Using quantifiable data allows for a more objective measure of identifying players who may follow a similar career trajectory.
-This approach offers a data-driven perspective on evaluating draft prospects, which adds to the subjective data provided by scouts.
-
-In this paper, we propose a methodology for calculating player similarity scores using a random forest model.
-We then apply this methodology to the quarterback position, focusing on how performance metrics from college football can be used to assess a player's potential at the professional level.
-By examining these scores, we provide a tool to improve the accuracy of draft projections and support more informed decision-making processes.
+The paper is organized as follows.
+In Section \ref{sec:data}, we describe a dataset on which we apply our methodology to evaluate NFL draft quarterback prospects, using ESPN's Total Quarterback Rating as the response variable and college statistics as the predictor variables.
+In Section \ref{sec:methods}, we derive similarity scores from a fitted random forest such that the model's predictions are equal to a similarity-weighted average of the training outcomes.
+In Section \ref{sec:results}, we interpret the predictions from the NFL quarterback draft prospect random forest model by ranking players from the training set according to their similarity scores induced by the adaptive nearest neighbors interpretation of random forests.
+In Section \ref{sec:software}, we describe the new R package rfcomp for calculating these similarity scores from fitted random forest objects.
+In Section \ref{sec:discussion}, we summarize the key findings.
 
 \subsection{Related Work}
-% Section 1.1 could use a little more narrative flow.
-You jump right in describing an article in the first sentence, but it would be helpful to provide more context.
-Is that article about sports?
-I think this section could be a little better organized into paragraphs with each paragraph representing one topic (e.g.
-interpreting machine learning algorithms, predicting player success in the draft, the concept of comps, etc.), with a sentence at the top of each paragraph explaining/introducing the topic.
-It would also be good to end this section with a paragraph explaining what your work accomplishes that hasn't been done before.
 
-Previous research has been done to attempt to introduce interpretability to random forest predictions.
+Previous research has been done to attempt to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
 \citet{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
 \citet{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models.
-In a sports context, \citet{metulini_measuring_2023} use generalized Shapely values to interpret individual player contributions based on a win probability model in the context of professional basketball.
+In a sports context, \citet{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.
 
 Draft models are fairly common in professional sports, as evaluating prospect talent can be key to success.
-One example of a draft model that uses QBR as the response variable is \citet{craig_predicting_2021}.
-They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season's metrics, scout grades, and player height.
+One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \citet{craig_predicting_2021}.
+They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season metrics, scout grades, and player height.
 Their predictions are easy to interpret directly from the logistic regression model.
-Another study uses their own metric of net points combined with games played to define NFL success.
-\citet{wolfson_quarterback_2011} use a combination of NFL Combine and college statistics to create a draft model.
-They use this model to analyze how effectively NFL teams leverage data available when drafting a quarterback.
-They use a logistic regression model, which results in predictions that are easily interpretable.
-\citet{berger_jumping_2021} uses principal component analysis (PCA) combined with a high order regression model to evaluate NBA prospects using their college statistics and Combine data.
-The reasoning behind the PCA predictions are unclear, resulting in a relatively uninterpretable draft model.
+\citet{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
+They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting a quarterback.
+Again, the logistic regression model results in predictions that are easily interpretable.
+\citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
+The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model.
 \citet{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics.
@@
 
-There are several articles from various sports reporters that discuss player comps from analysts.
-Pro Football Focus published their comps for their top 20 prospects this year, however the only quarterback prospect on the list was Cam Ward, whom they compared to Ben Roethlisberger \citep{sikkema_2025_2025}.
-The Giants also published an article with various analyst opinions on top prospect comps.
-This article included three quarterbacks: Cam Ward, Shedeur Sanders, and Jalen Milroe \citep{noauthor_draft_2025}.
-A final example is an article from CBS Sports that describes current prospect comps to past NFL busts \citep{trapasso_nfl_2025}.
-
-While the articles discussed show the previous work that has been done to interpret machine learning models, evaluate prospects, and make player comparisons, none of them have combined these three concepts.
-This paper introduces interpretability to random forest predictions in the context of sports, specifically American Football.
-In doing so, we also introduce a method of calculating quantifiable player comps between the training players in the dataset and the prospects.
+While past work in sport analytics has attempted to predict draft prospect outcomes using interpretable methods (i.e.
+logistic regression) or machine learning methods with greater predictive accuracy, the novelty of the current work is that we achieve interpretability with a machine learning moethod.
+In particular, we tie our novel interpretation of random forest predictions to the concept of comps, which is well understood by scouts and executives in the sports industry.
+The implementation of our method is available in the treecomp R package, allowing practitioners to easily apply it to their own fitted random forests.
 
 \section{Data}
+\label{sec:data}
 
@@
 \section{Methods}
+\label{sec:methods}
 
@@
 \subsubsection{Similarity Formula}
+
 Using the random forest model's relationship to \textit{k}-NN, we can derive a formula for similarity scores between testing player \textit{i} and training player \textit{j}.
@@
 \section{Results}
+\label{sec:results}
 
@@
 
-\input{../../tables/top_ten.tex}
+\input{../../tables/top_ten}
 
@@
 \resizebox{\textwidth}{!}{
-  \input{../../tables/side_by_side_similarity.tex}}
+  \input{../../tables/side_by_side_similarity}}
   \caption{\textit{A table of the top similarity percentages for each of the four selected prospects.
@@
 
+\section{Software}
+\label{sec:software}
+
 \section{Discussion}
+\label{sec:discussion}
 

### 2025-05-19 articles/arxiv/paper.tex
fine-tune data section

@@
 
-There were two sources for our data.
-The NFL data for ESPN's Total Quarterback Rating (QBR) came from \href{https://pro-football-reference.com}{pro-football-reference.com}.
-For the corresponding college statistics, we used data from \href{https://sports-reference.com}{sports-reference.com}.
-The college statistics only include schools that are a part of the NCAA Division 1 Football Bowl Subdivision (FBS).
-We were able to obtain college statistics dating back to 1984.
-However, we were constrained by the ESPN QBR metric, which has calculations starting at the 2006 season.
+All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}.
+Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who averaged more than six games played per season and more than five passing attempts per season.
+For each of these quarterbacks, we obtained the following data from their college career: games played per season, passing attempts per season, passing yards per season, passing yards per attempt, passing touchdowns per season, interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season conference, final-season games played, final-season passing yards per attempt, and final-season passer rating.
 
-The data from Sports Reference contained player passing statistics for each season.
-It contained categories for player, team, conference, games, completions, attempts, yards, touchdowns, touchdown percentage, interceptions, interception percentage, yards per attempt, adjusted yards per attempt, yards per completion, yards per game, passing efficiency rating, awards, and season.
-In addition to the passing statistics, we included player rushing statistics for each season from Sports Reference.
+In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) \citep{burke_how_2016} from each season between 2006 and 2024 (inclusive).
+Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play.
+It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time.
+The range of QBR is from 0 to 100, and \citet{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
+For each quarterback, we calculate their average QBR per season, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero.
+Here the scale of QBR is helpful because players who reach the NFL cannot be rated below players who never reach the NFL.
 
-We obtained NFL season passing statistics from Pro Football Reference for the 2006-2024 seasons.
-The resulting data frame contained columns for 28 different passing metrics.
-However, we only used QBR and overall pick from this model.
-
-\subsection{ESPN's Total Quarterback Rating}
-
-The NFL response variable we used in this paper is ESPN's Total Quarterback Rating (QBR).
-It is an EPA-based metric that is adjusted for home-field advantage, defensive strength, and ``garbage time'' and then scaled from 0-100 \citep{noauthor_how_2016}.
-QBR provides a more holistic evaluation of quarterbacks in the NFL while also considering other factors affecting performance.
-Additionally, a 0-100 scale is important in this context because it offers an interpretable standard frame of reference.
-It also allows us to assign a meaningful value to quarterbacks who never played in the NFL, compared to raw EPA (where the lack of such scale results in negative values for some players).
-For example, it would not make much sense for a quarterback who receives playing time in the NFL to have a lower value than a quarterback who never made it to the NFL, but this would happen if using raw EPA as the response variable.
-Using QBR as the response variable addresses this issue while incorporating other important aspects of game-time performance.
+Average QBR per season is the response variable describing the success of quarterback prospects.
+Because of the date ranges of the data, we do not calculate this perfectly for all of the quarterbacks in our dataset.
+On one end of the extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
+These quarterbacks have a QBR of zero in our dataset.
+On the other end of the extremes, quarterbacks who finished their college careers in 2023 and have not yet appeared in the NFL may yet do so.
+These quarterbacks, too, have a QBR of zero in our dataset.
+This is a limitation of the data.
 

### 2025-05-19 articles/arxiv/paper.tex
very minor wording change

@@
 Because of the date ranges of the data, we do not calculate this perfectly for all of the quarterbacks in our dataset.
-On one end of the extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
+On one extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
 These quarterbacks have a QBR of zero in our dataset.
-On the other end of the extremes, quarterbacks who finished their college careers in 2023 and have not yet appeared in the NFL may yet do so.
+On the other extreme, quarterbacks who finished their college careers in 2023 and have not yet appeared in the NFL may yet do so.
 These quarterbacks, too, have a QBR of zero in our dataset.

### 2025-05-20 articles/arxiv/paper.tex
fine-tune methods section

@@
 
-\usepackage{authblk, array, float, graphicx, hyperref, natbib, subcaption, amsmath}
+\usepackage{amsmath, amsfonts, array, authblk, float, graphicx, hyperref, natbib, subcaption}
 \usepackage[margin=1in]{geometry}
@@
 All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}.
-Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who averaged more than six games played per season and more than five passing attempts per season.
-For each of these quarterbacks, we obtained the following data from their college career: games played per season, passing attempts per season, passing yards per season, passing yards per attempt, passing touchdowns per season, interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season conference, final-season games played, final-season passing yards per attempt, and final-season passer rating.
+Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
+The data for each player-season include team, conference, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
 
@@
 
-\subsection{Data Preparation}
+\subsection{Random Forest}
 
-\subsubsection{Data Cleaning}
+Random forests are a machine learning method typically used for classification or regression \citep{breiman_random_2001}.
+The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision within each bootstrap sample, randomly choosing at each node a subset of the features to consider for the split.
+The random forest's prediction is the average of the predictions of the individual treees.
+By fitting multiple decision trees, random forests generally achieve much lower-variance predictions than individual decision trees.
+The randomness introduced by the bootstrap resampling of the data and by randomly limiting eligible features at each split drives down the correlation between the trees in the forest, further reducing the variance of the fitted values.
 
-In order to clean the data frame containing college statistics, we started by joining the rushing dataframe with the passing dataframe by player and season.
-We transformed the data frame to contain only one player per row by taking the average of their games played, completions, attempts, passing yards, passing passing touchdowns, interceptions, rushing attempts, rushing yards, and rushing touchdowns.
-In addition to career statistics, we added columns for the career games played, career seasons played, career yards per attempt, final season's yards per attempt, final season, final season's games played, final season's passer rating, final season's college, and the final season's conference.
-Next, we created a binary variable for winning the Heisman Award.
-We also created a variable for how each player finished in the Heisman voting in their final college season and how many years the player was an All-American selection.
-Finally, we filtered the dataset to remove players who had less than 5 average attempts per season and players with less than less than 6 career games and whose last season was before 2024 (as not to include players still in college).
+We trained the random forest model using the R package ranger \citep{wright_ranger_2015}, which is a fast implementation of the random forest algorithm.
+The response variable was average QBR per season (zero for players who never appeared in the NFL).
+The college statistics used as predictor variables were games played per season, completed passes per season, attempted passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season conference, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
 
-In the NFL dataset, we calculated the average QBR per season played for each player.
-The resulting data frame contained one row per player with columns for player name and average QBR.
-We combined this data frame with the college data frame and replaced any missing QBR values (due to never having played in the NFL) with zero.
-We also filtered the data to remove any players whose final year in college was before 2000 since our NFL data only dates back to the 2006 season.
-Our final dataset included statistics for 2,115 quarterbacks.
-
-\subsection{Random Forest Model}
-
-Random forests are a machine learning method typically for classification or regression.
-For regression tasks, \textit{n} decision trees are created during training.
-Each tree produces a prediction of the response variable, and the final prediction is the average of the individual trees' predictions.
-The use of multiple random decision trees reduces overfitting by introducing more bias, which reduces variance.
-
-We trained the random forest model using the R package ranger, which is a fast implementation of the random forest algorithm \citep{wright_ranger_2015}.
-The response variable was average QBR per season (zero for players who never appeared in the NFL).
-The college statistics used as predictor variables were career games played, average completions per season, average attempts per season, average yards per season, average touchdowns per season, average interceptions per season, average rushing attempts per season, average rushing yards per season, average rushing touchdowns per season, career yards per attempt, career All-American designations, final season yards per attempt, final season games played, final season passer efficiency rating, final season Heisman award voting, whether the player won the Heisman award, and final season NCAA conference.
-We implemented a random 70\%/30\% training/testing data split, and the random forest contained 500 decision trees.
-
-To determine the best hyperparameters, we used the tuneRanger function from the tuneRanger package \citep{probst_tuneranger_2018} with 100 iterations and RMSE as the measure to optimize.
-We found the best parameters to be \texttt{min.node.size} = 41 and \texttt{mtry} = 4.
+To train the random forest model, we first implemented a random 70\%/30\% train/test split.
+We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement.
+To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
+After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be X and the optimal minimum node size to be XX.
 
@@
 
-\subsubsection{\textit{k}-Nearest Neighbors}
+\citet{lin_random_2006} previously established the relationship between $k$-NN and the random forest algorithm.
+In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data.
+We draw $B$ bootstrap samples of the data indexed by $b \in \{1, ..., B\}$, and we fit a decision tree on each of these samples.
+We use $n_{b,i}$ to denote the number of times observation $i$ is included in bootstrap sample $b$, and we use $\mathcal{T}_{b,i}$ to denote the terminal node of observation $i$ in decision tree $b$.
+We use $|\mathcal{T}_{b,i}|$ to denote the number of training observations (with repetition) which belong to $\mathcal{T}_{b,i}$.
+If observation $i$ is not included in bootstrap sample $b$, then $n_{b,i} = 0$, and we define $\mathcal{T}_{b,i} = \{i\}$ and $|\mathcal{T}_{b,i}| = 1$.
 
-\citet{lin_random_2006} previously established that there is a relationship between \textit{k}-nearest neighbors and random forest.
-Given that the training data is $\{(x_i,y_i),i=1, ...,n\}$, a prediction for $x_0$ from any given tree in a random forest model is $$\hat{y}_m(x_0)=\sum_{i=1}^{n}W_{im}y_i$$
-where:
-\[W_{im} = 
-\begin{cases}
-\frac{1}{k_m}, & \text{if } x_i \text{ is in the same terminal node as } x_0 \text{ in tree } m, \\
-0, & \text{otherwise}.
-\end{cases} 
-\]
-and $k_m$ denotes the number of training points in that terminal node.
+From this setup, if we query the model at a new feature vector $\vec x_0 \in \mathbb{R}^p$, then the prediction of decision tree $b$ is given by
+$$
+  \hat{y}_b(\vec x_0) = \sum_{i = 1}^{n} w_{b,0,i} \cdot y_i,
+$$
+where $w_{b,0,i} = n_{b,i} \cdot \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ which comes from observation $i$.
+The weights $w_{b,\cdot,0}$ can be thought of as defining a local neighborhood around $\vec x_0$ for tree $b$.
 
-This setup defines a local neighborhood around $x_0$ for tree $m$.
+The prediction of the random forest for the feature vector $\vec x_0$ is simply the average of the predictions of the individual trees:
+\begin{equation*}
+  \hat{y}(\vec x_0) = \frac{1}{B} \sum_{b = 1}^B \left(\sum_{i = 1}^{n} w_{b,0,i} \cdot y_i\right),
+\end{equation*}
+which can be rewritten as
+\begin{equation}
+  \label{eqn:rf-as-knn}
+  \hat{y}(\vec x_0) = \sum_{i = 1}^n \bar w_{0,i} \cdot y_i,
+\end{equation}
+where $\bar w_{0,i} = \left(\sum_{b = 1}^B w_{b,0,i}\right) / B$ is the average weight across the trees.
 
-The final random forest prediction results from averaging the individual tree predictions $$\hat{y}(x_0)=\sum_{i=1}^M\bar{W_i}y_i$$
-where the average weight, $\bar{W_i}$, is defined as $$\frac{1}{M}\sum_{m=1}^MW_{im},$$ where $M$ is the total number of trees in the forest.
+This formulation (\ref{eqn:rf-as-knn}) demonstrates how the random forest can be interpreted as a form of \textit{k}-NN, where the forest itself is a weighted neighborhood scheme and neighborhoods are defined by tree structure rather than distance.
+The training observations that share a leaf node with $\vec x_0$ in tree $b$ are considered its neighbors.
+The final prediction results from averaging the results across all trees, where training points that show up more frequently as neighbors receive greater weight.
 
-This formulation shows that a random forest can be interpreted as a form of \textit{k}-NN, where the forest itself is a weighted neighborhood scheme and neighborhoods are defined by tree structure rather than distance.
-The training points that share a leaf node with $x_0$ in tree $m$ are considered its neighbors.
-The final prediction results from averaging the results across all trees, where training points that show up as neighbors more frequently receive higher weight.
+We take this a step further and emphasize the interpretation of $\bar w_{0,i}$ as the similarity of $\vec x_0$ to $\vec x_i$. Explicitly, this {\it similarity score} is defined as
+\begin{equation}
+  \bar w_{0,i} = \frac{1}{B}\sum_{b = 1}^B
+    \frac{n_{b,i}}{|\mathcal{T}_{b,0}|} \cdot \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\}.
+\end{equation}
 
-\subsubsection{Similarity Formula}
-
-Using the random forest model's relationship to \textit{k}-NN, we can derive a formula for similarity scores between testing player \textit{i} and training player \textit{j}.
-We derived the formula for the similarity score between player $i$ and player $j$ to be
-$$S(i,j) = \frac{1}{M}\sum_{k=1}^{M}{\frac{N_{j,T_{i,k}}}{|T_{i,k}|}}$$
-where:
-\begin{itemize}
-    \item $|T_{i,k}|$ is the total number of players in player \textit{i}'s leaf nodes of tree \textit{k} 
-    \item $N_{j,T_{i,k}}$ is the number of times player \textit{j} appears in player \textit{i}'s leaf nodes in tree \textit{k}
-\end{itemize}
-
-We can use these similarity scores to recreate the predictions for player $i$ from the random forest by taking the weighted average of the training players' observed performance using the similarity scores as the weight.
-Note that for each tree $k$, the weight $W_{jk}$ is $\frac{1}{|T_{i,k}|}$ if $j \in T_{i,k}$ and $0$ otherwise.
-Therefore, player $j$'s contribution across all trees can be defined as
-
-$$\bar{W}_j = \frac{1}{M} \sum_{k=1}^M \frac{N_{j,T_{i,k}}}{|T_{i,k}|} = S(i,j),$$
-
-and substituting back into the prediction formula, we obtain:
-\[
-\hat{y}_i = \sum_{j=1}^n \bar{W}_j y_j = \sum_{j=1}^n S(i,j) y_j.
-\]
-The similarity score $S(i,j)$ serves as the prediction weight for training player $j$ when predicting the output for test player $i$.
+Note that the similarity score is not symmetric.
+In general, $\bar w_{0,i} \ne \bar w_{i,0}$.
+In fact, as we have defined it, $\bar w_{i,0}$ is undefined if $\vec x_0$ is not in the training set.
+The similarity score $\bar w_{0,i}$ can be thought of as describing how similar is observation 0 to observation $i$, with similarity determined adaptively in a meaningful way so as to optimize random forest predictions.
+In contrast, defining similarity based on Euclidean distance would not put more weight on features which are more important for predicting outcomes.
+The similarity score $\bar w_{0,i}$ comes with the nice property that the prediction for $\vec x_0$ is recovered by calculating the weighted average of the outcome values in the training data, using $\bar w_{0,i}$ as the weights.
 
@@
 
-These four quarterbacks were selected to illustrate the model’s interpretability and its alignment with real-world outcomes.
+These four quarterbacks were selected to illustrate the model's interpretability and its alignment with real-world outcomes.
 We used Shedeur Sanders and Cam Ward as two examples because they were both initially expected to go in the first round of this year's draft.

### 2025-05-20 random-forest-comps.R
modify the tuning method

@@ -81,7 +81,7 @@ train_data <- QBR_passing_data_combined[train_indices, ]
 test_data <- QBR_passing_data_combined[-train_indices, ]
 
 # Model tuning
-task_data <- QBR_passing_data_combined %>%
+task_data <- train_data %>%
   select(c_career_games, c_career_cmp, c_career_att, c_career_yds, c_career_td, c_career_int, last_games, last_passer_rating, AA, last_H_vote, won_H, last_conference, mean_QBR, yds_per_att, final_yds_per_att, c_rush_att, c_rush_yds, c_rush_td)
 task_data$last_conference <- as.factor(task_data$last_conference)
 task <- makeRegrTask(data = task_data, target = "mean_QBR")
@@ -91,7 +91,8 @@ tuned_model <- tuneRanger(
   task,
   num.trees = 500, 
   iters = 100,
-  tune.parameters = c("mtry", "min.node.size"))
+  tune.parameters = c("mtry", "min.node.size"),
+  parameters = list(replace = TRUE))
 
 print(tuned_model)
 
@@ -102,7 +103,7 @@ rf_model <- ranger(mean_QBR ~ c_career_games + c_career_cmp + c_career_att + c_c
                    + final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
                    data = train_data,  
                    num.trees = 500, importance = "impurity",
-                   keep.inbag = TRUE, min.node.size = 41, mtry = 4)
+                   keep.inbag = TRUE, min.node.size = 84, mtry = 10)
 
 test_terminal_nodes <- predict(rf_model, data = test_data, type = "terminalNodes")$predictions
 

### 2025-05-30 scripts/download_data.R
re-organize data and figures

@@ -23,7 +23,7 @@ for (year in 1984:2024) {
 }
 
 head(passing_data)
-write.csv(passing_data, "cfb_passing_stats_1984_2024.csv", row.names = TRUE)
+write.csv(passing_data, "data/cfb_passing_stats_1984_2024.csv", row.names = TRUE)
 
 # Construct college rushing dataframe
 rushing_data <- data.frame()
@@ -51,7 +51,7 @@ for (year in 2000:2024) {
 }
 
 head(rushing_data)
-write.csv(rushing_data, "cfb_rushing_stats_2000_2024.csv", row.names = TRUE)
+write.csv(rushing_data, "data/cfb_rushing_stats_2000_2024.csv", row.names = TRUE)
 
 # Construct NFL QBR dataframe
 QBR_passing_data <- data.frame()
@@ -75,4 +75,4 @@ for (year in 2006:2024) {
 }
 
 head(QBR_passing_data)
-write.csv(QBR_passing_data, "QBR_stats_2006_2024.csv", row.names = TRUE)
+write.csv(QBR_passing_data, "data/QBR_stats_2006_2024.csv", row.names = TRUE)

### 2025-05-30 scripts/generate_results.R

@@ -6,9 +6,9 @@ library(mlr)
 library(ggplot2)
 library(xtable)
 
-cfb_data <- read.csv("cfb_passing_stats_1984_2024.csv") 
+cfb_data <- read.csv("data/cfb_passing_stats_1984_2024.csv") 
 
-cfb_rush <- read.csv("cfb_rushing_stats_2000_2024.csv") %>%
+cfb_rush <- read.csv("data/cfb_rushing_stats_2000_2024.csv") %>%
   mutate(Player = str_replace(Player, "\\*", "")) %>%
   select(Player, Season, Team, rush_att = Att, rush_yds = Yds, rush_td = TD)
 
@@ -59,7 +59,7 @@ cfb_career_data <- cfb_data %>%
          c_career_tot_games > 6,
          c_career_att > 5)
 
-QBR_passing_data <- read.csv("QBR_stats_2006_2024.csv") %>%
+QBR_passing_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
   filter(Pos == "QB") %>%
   group_by(Player) %>%
   summarise(mean_QBR = mean(QBR),
@@ -125,34 +125,49 @@ importance_df <- data.frame(
   Importance = importance_scores
 )
 
-ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
-  geom_bar(stat = "identity") +
-  coord_flip() +  
-  labs(title = "Variable Importance",
-       x = "", y = "Importance") +
-  theme_minimal()
+{
+  sputil::open_device("figures/variable_importance.pdf", height = 5)
+  plot <- ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
+    geom_bar(stat = "identity") +
+    coord_flip() +  
+    labs(title = "Variable Importance",
+         x = "", y = "Importance") +
+    theme_minimal()
+  print(plot)
+  dev.off()
+}
 
 new_plot_data <- QBR_passing_data_combined %>%
   mutate(predictions = predict(rf_model, data = QBR_passing_data_combined)$predictions)
 
-ggplot(new_plot_data, aes(x = predictions, y = mean_QBR)) +
-  geom_point() +
-  labs(title = "Predictions vs. Actuals",
-       x = "Predicted QBR",
-       y = "Actual QBR") + 
-  geom_abline(intercept=0, slope=1) +
-  xlim(0,100) +
-  ylim(0,100) +
-  theme_minimal()
[…]

### 2025-05-30 articles/arxiv/paper.tex
fine-tine Section 4.1

@@
 Average QBR per season is the response variable describing the success of quarterback prospects.
-Because of the date ranges of the data, we do not calculate this perfectly for all of the quarterbacks in our dataset.
+In the dataset, 91\% of college quarterbacks achieve zero QBR in the NFL, so the distribution is very right-skewed.
+Because of the date ranges of the data, we do not calculate QBR perfectly for all of the quarterbacks in our dataset.
 On one extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
@@
 
-\subsection{Model Performance}
+\subsection{Random Forest}
 
-The model's test root mean squared error (RMSE) is 9.88, and 33.87\% of the model's variance is explained. A plot of the variable importance is below.
+Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
+The model's test RMSE was 9.89 units of QBR, corresponding to explaining 35.1\% of the variance of QBR in the test set.
+Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success.
+Note that these three variables are highly correlated with each other---each pairwise correlation is at least 0.94.
+The next most important variables are final-season Heisman voting, average passing attempts per season and final-season passer rating.
 
@@
     \centering
-%    \includegraphics[width=0.75\linewidth]{model plots/variable_importance.png}
-    \caption{\textit{A histogram showing the importance of each predictor variable in the random forest model predicting career average QBR.}}
+    \includegraphics[width=0.75\linewidth]{../../figures/variable_importance.pdf}
+    \caption{\textit{Variable importance plot for random forest trained to predict NFL QBR for college quarterbacks.}}
     \label{fig:importance}
@@
 
+Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted QBR against career passing yards per season and final-season passer rating.
+These are two of the most important predictors in the model, and they have a relatively low correlation.
+The extreme outlier by passing yards per season is Bailey Zappe, who played only one FBS season at Western Kentucky in Conference USA and was selected by the New England Patriots in the fourth round of the 2022 NFL Draft.
+The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed a QBR of 40.
+This speaks to the large number of college quarterbacks who achieve zero QBR in the NFL and the difficulty of distinguishing between prospects based on college performance statistics alone.
+
 \begin{figure}[H]
     \centering
-%    \includegraphics[width=0.75\linewidth]{model plots/3d_plot.png}
-    \caption{\textit{This plot shows the predictions on the testing set as a function of college touchdowns per season and final season passer rating.
-The color of the point represents the QBR prediction value on a continuous scale, while the x-axis represents college yards per season, and the y-axis represents the final season passer rating.}}
+    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
+    \caption{\textit{Predicted QBR plotted against average passing yards per season and final-season passer rating.}}
+    \label{fig:3d-plot}
 \end{figure}
 
-Figure \ref{fig:predicted-vs-actuals} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
+Figure \ref{fig:predicted-vs-actuals} shows actual vs.
+predicted QBR for all quarterbacks in the dataset.
+Recall that 91\% of actual QBRs are zero, which the visualization does not make clear.
+The errors are very right-skewed, and this illustrates why no college quarterback exceed a QBR prediction of 50.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
+  \caption{\textit{Predicted (from random forest) and actual QBR for all quarterbacks in the dataset.
+The vats majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}}
+  \label{fig:predicted-vs-actuals}
+\end{figure}
+
+\subsection{Case Study: 2025 NFL Draft}
+
+Table \ref{tab:top-ten} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
 Cam Ward is predicted to have the highest QBR, which matches the common consensus that he is the best quarterback in this draft class and the fact he was taken first overall in the draft.
@@
 
-\begin{figure}[H]
+\begin{table}[H]
   \centering
-%    \includegraphics[width=0.75\linewidth]{model plots/predicted_vs_actuals.png}
-    \caption{\textit{Actual QBR plotted against the random forest generated QBR prediction for the testing dataset.}}
-    \label{fig:predicted-vs-actuals}
-\end{figure}
-
   \input{../../tables/top_ten}
-
-\subsection{2025 NFL Draft Comps}
+  \caption{\textit{The top ten career average QBR predictions for this year's draft prospects.
+This only includes players who are eligible to be drafted this year in the NFL and who played at an NCAA Division 1 Football Bowl Subdivision school during the 2024 season.
+The predictions were produced from a random forest model that used the full dataset (as opposed to just the training dataset.}}
+  \label{tab:top-ten}
+\end{table}
 
@@
     \centering
-%    \includegraphics[width=0.24\linewidth]{prospect plots/cw_plot1.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/ss_plot1.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/dg_plot1.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/jd_plot1.png} \\
-%    \includegraphics[width=0.24\linewidth]{prospect plots/cw_plot2.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/ss_plot2.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/dg_plot2.png}
-%    \includegraphics[width=0.24\linewidth]{prospect plots/jd_plot2.png}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_dart.pdf} \\
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
     \caption{\textit{Top: Histogram of training players’ QBR values weighted by the proportion of shared nodes with the prospect (similarity).

### 2025-05-30 scripts/generate_results.R
update color gradient and plotting range for 3d_plot

@@ -142,26 +142,32 @@ new_plot_data <- QBR_passing_data_combined %>%
 
 {
   sputil::open_device("figures/3d_plot.pdf", height = 5)
-  plot <- ggplot(test_data, aes(x = c_career_td, y = last_passer_rating, color = predictions)) +
-    geom_point(size = 1) +  # Use points to show individual data points
-    scale_color_gradient(low = "blue", high = "red") +  # Choose your color gradient
-    labs(title = "Predictions from Random Forest Model",
-         x = "College Touchdowns per Season",
-         y = "Final Season Passer Rating",
-         color = "Predictions") +
-    theme_minimal()
+  plot <- new_plot_data |>
+    ggplot(aes(x = c_career_yds, y = last_passer_rating, color = predictions)) +
+    geom_point(size = 1) +
+    scale_color_viridis_c() +
+    labs(
+      x = "Passing Yards per Season (College)",
+      y = "Final-Season Passer Rating (College)",
+      color = "Predicted QBR"
+    ) +
+    coord_cartesian(xlim = c(0, 6000), ylim = c(0, 250)) +
+    theme_minimal() +
+    theme(legend.position = "inside", legend.position.inside = c(0.85, 0.25))
   print(plot)
   dev.off()
 }
 
 {
-  sputil::open_device("figures/predicted_vs_actuals.pdf", height = 5)
+  sputil::open_device("figures/predicted_vs_actuals.pdf", height = 5, width = 5)
   plot <- ggplot(new_plot_data, aes(x = predictions, y = mean_QBR)) +
     geom_point() +
-    labs(title = "Predictions vs. Actuals",
-         x = "Predicted QBR",
-         y = "Actual QBR") + 
-    geom_abline(intercept=0, slope=1) +
+    labs(
+      title = "Predictions vs. Actuals",
+      x = "Predicted QBR",
+      y = "Actual QBR"
+    ) + 
+    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
     xlim(0,100) +
     ylim(0,100) +
     theme_minimal()

### 2025-05-30 scripts/generate_results.R
update the second Ranger model with the new hyperparameters

@@ -181,7 +181,7 @@ full_rf_model <- ranger(mean_QBR ~ c_career_games + c_career_cmp + c_career_att 
                         + final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
                         data = QBR_passing_data_combined,  
                         num.trees = 500, importance = "impurity",
-                        keep.inbag = TRUE, min.node.size = 41, mtry = 4)
+                        keep.inbag = TRUE, min.node.size = 84, mtry = 10)
 
 cfb_career_data_current <- cfb_data %>%
   filter(last_season == 2024) %>%

### 2025-06-04 scripts/generate_results.R
fine-tune Section 4.2

@@ -188,22 +188,16 @@ cfb_career_data_current <- cfb_data %>%
   mutate_all(~ replace(., is.na(.), 0))
 
 cfb_career_data_current$predictions <- round(predict(full_rf_model, data = cfb_career_data_current)$predictions, 3)
-player_predictions_table <- cbind(cfb_career_data_current$Player, cfb_career_data_current$predictions)
 
-player_predictions_table <- as.data.frame(player_predictions_table)
+cfb_career_data_current %>%
+  select(Player, predictions) %>%
+  # Remove draft-ineligible players
+  filter(!Player %in% c("Darian Mensah", "Cade Klubnik", "E.J. Warner")) %>%
+  arrange(-predictions) %>%
+  head(10) %>%
+  mutate(predictions = sprintf("%.1f", predictions)) %>%
+  sputil::write_latex_table(file = "tables/top_ten.tex")
 
-colnames(player_predictions_table) <- c("Player", "Prediction")
-player_predictions_table <- player_predictions_table %>% 
-  arrange(desc(as.numeric(Prediction))) %>%
-  filter(Player != "Darian Mensah",
-         Player != "Cade Klubnik",
-         Player != "E.J. Warner")
-
-top_10 <- head(player_predictions_table, 10)
-
-colnames(top_10) <- c("Player", "Predicted QBR")
-
-latex_table <- xtable(top_10)
 
 test_names <- cfb_career_data_current$Player
 test_terminal_nodes <- predict(full_rf_model, data = cfb_career_data_current, type = "terminalNodes")$predictions
@@ -267,13 +261,12 @@ ss_plot_data <- data.frame(
   sputil::open_device("figures/prospect_histogram_sanders.pdf", height = 3, width = 3)
   plot <- ggplot(ss_plot_data, aes(x = QBR, y = ..density.., weight = sim_score)) +
     geom_histogram(binwidth = 5, color = "blue", fill = "blue", alpha = 0.8) +
-    geom_density(aes(x = QBR, y = ..density..), 
-                 color = "red", fill = "red", alpha = 0.5, adjust = 0.5) +
     theme_minimal() +
     labs(x = "QBR", y = "Weights",
          title = "Shedeur Sanders") +
     geom_vline(aes(xintercept = cfb_career_data_current[ss_index, ]$predictions), 
[…]

### 2025-06-04 data_scraping.R
Update data_scraping.R

@@ -76,3 +76,31 @@ for (year in 2006:2024) {
 
 head(QBR_passing_data)
 write.csv(QBR_passing_data, "QBR_stats_2006_2024.csv", row.names = TRUE)
+
+schedule_strength <- data.frame()
+
+for (year in 2000:2024) {
+  url <- paste0("https://www.sports-reference.com/cfb/years/", 
+                year, "-standings.html")
+  
+  page <- read_html(url)
+  
+  raw_data <- page %>%
+    html_node("table") %>% 
+    html_table(header = FALSE, fill = TRUE) 
+  
+  colnames(raw_data) <- raw_data[2, ] 
+  data <- raw_data[-c(1, 2), c(2, 13)]
+  
+  data$Season <- year
+  
+  schedule_strength <- bind_rows(schedule_strength, data)
+  
+  print(year)
+  
+  Sys.sleep(runif(6, 8, 10))
+}
+
+head(schedule_strength)
+write.csv(schedule_strength, "cfb_schedule_strength_2000_2024.csv", row.names = TRUE)
+

### 2025-06-04 scripts/download_data.R
update to include the strength of schedule df

@@ -76,3 +76,30 @@ for (year in 2006:2024) {
 
 head(QBR_passing_data)
 write.csv(QBR_passing_data, "data/QBR_stats_2006_2024.csv", row.names = TRUE)
+
+schedule_strength <- data.frame()
+
+for (year in 2000:2024) {
+  url <- paste0("https://www.sports-reference.com/cfb/years/", 
+                year, "-standings.html")
+  
+  page <- read_html(url)
+  
+  raw_data <- page %>%
+    html_node("table") %>% 
+    html_table(header = FALSE, fill = TRUE) 
+  
+  colnames(raw_data) <- raw_data[2, ] 
+  data <- raw_data[-c(1, 2), c(2, 13)]
+  
+  data$Season <- year
+  
+  schedule_strength <- bind_rows(schedule_strength, data)
+  
+  print(year)
+  
+  Sys.sleep(runif(6, 8, 10))
+}
+
+head(schedule_strength)
+write.csv(schedule_strength, "cfb_schedule_strength_2000_2024.csv", row.names = TRUE)

### 2025-06-04 articles/arxiv/paper.tex
minor changes to table caption

@@
 
-\title{A comps-based approach for interpreting tree-based predictions}
+\title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}
 \author[1]{Elisabeth Millington}
@@
 \begin{equation}
+  \label{eqn:similarity}
   \bar w_{0,i} = \frac{1}{B}\sum_{b = 1}^B
@@
 Table \ref{tab:top-ten} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
-Cam Ward is predicted to have the highest QBR, which matches the common consensus that he is the best quarterback in this draft class and the fact he was taken first overall in the draft.
-Dillon Gabriel and Shedeur Sanders were predicted second and third respectively.
-While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he has had a very consistent college career and finished the highest out of any quarterback in Heisman voting this year.
+Cam Ward is predicted to have the highest QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft.
+Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest QBRs, respectively.
+While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025.
+Jackson Dart's draft stock increased substantially after the Scouting Combine and his pro day.
 
@@
   \centering
-  \input{../../tables/top_ten}
-  \caption{\textit{The top ten career average QBR predictions for this year's draft prospects.
-This only includes players who are eligible to be drafted this year in the NFL and who played at an NCAA Division 1 Football Bowl Subdivision school during the 2024 season.
-The predictions were produced from a random forest model that used the full dataset (as opposed to just the training dataset.}}
+  \begin{tabular}{l|r}
+    Player & Predicted QBR\\
+    \hline
+    \input{../../tables/top_ten.tex}
+  \end{tabular}
+  \caption{\textit{Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class.
+This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
   \label{tab:top-ten}
@@
 
-Table 2 below shows the result of calculating similarity scores for the four chosen quarterback prospects.
-These scores are essentially the percentage of the prospect's leaf nodes that are shared with that training player.
-We can use these to interpret the predictions from Table 1 as the weights each of the training players has on the prospect's QBR prediction.
-We can see that for Cam Ward, Shedeur Sanders, and Dillon Gabriel, the vast majority of the training players with the top ten similarity scores are relatively successful starting (or formerly starting) quarterbacks.
-Intuitively, since they had the three highest predictions for this draft class, it would make sense that their predictions are based on players who are/were successful in the NFL (i.e.
-had a high career average QBR).
-For Jaxson Dart, we can see that only four of the ten players who have the highest similarity scores were ever starters in the NFL.
-This means that the other six players likely have a QBR value that is close to zero, which explains why his prediction is significantly lower than the other three prospects.
+The predictions are interesting, but the focus of the present work is interpreting the predictions.
+Table \ref{tab:side-by-side-similarity} reports the ten most similar historical draft prospects for the top four quaterback prospects: Cam Ward, Shedeur Sanders, Jaxson Dart and Dillon Gabriel.
+Recall that the similarity scores reported in this table are interpretable as the weight of each historical prospect in the predictions in Table \ref{tab:top-ten}.
+Football fans will notice many household names among the comps for Ward, Sanders and Gabriel, but the comps for Dart are less impressive.
+Of note, the highest similarity score for any of these prospects is only 3.0\%.
+This is a helpful reminder that quarterback career outcomes are noisy, and the random forest model learns to spread out the weight across historical players, rather than concentrating a lot of weight in the single most similar player.
 
-These four quarterbacks were selected to illustrate the model's interpretability and its alignment with real-world outcomes.
-We used Shedeur Sanders and Cam Ward as two examples because they were both initially expected to go in the first round of this year's draft.
-Shedeur Sanders' draft position did unexpectedly slide, and he ended up being taken in the fifth round in the draft.
+These four quarterback prospects illustrate the model's interpretability and its alignment with real-world outcomes.
+Sanders and Ward were both initially expected to go in the first round of the 2025 draft.
+Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft.
 However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance \citep{mckenna_what_nodate}.
-We chose Jaxson Dart because his draft stock continuously increased after the NFL Scouting Combine and his Pro Day, and he was one of the two quarterbacks taken in the first round of the NFL draft.
-Finally, we chose Dillon Gabriel because he finished the highest of all quarterbacks in Heisman voting this year (3rd overall).
-We used a model containing the full dataset (not just the training data) to calculate the similarity scores for each of these four players using the above-described methodology.
-It is worth noting that both Cam Ward and Shedeur Sanders started their college careers at teams that are not in the NCAA D1 FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State.
-This means that their college statistics used in the model do not include their statistics from these schools.
-The table below serves as a tool for interpreting the above random forest predictions for these draft prospects.
-The percentages offer a quantifiable measure of similarity based on college statistics that underlie the random forest prediction output, allowing us to interpret the predictions better.
-We can see that Cam Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
-Dillon Gabriel has a slightly higher QBR prediction than Shedeur Sanders, even though Shedeur was considered a better prospect.
-However, Gabriel finished higher than Shedeur in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft.
-Jaxson Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration.
-He also performed well in his interviews with teams, which is not a quantifiable measure.
+In the end, Dart was the other quarterback taken in the first round of the draft.
 
@@
   \resizebox{\textwidth}{!}{
-  \input{../../tables/side_by_side_similarity}}
-  \caption{\textit{A table of the top similarity percentages for each of the four selected prospects.
-The similarity percentage is the percentage of nodes that the training quarterback and prospect ended in the same node based on the predictor variables.}}
+    \begin{tabular}{lr|lr|lr|lr}
+      \multicolumn{2}{c|}{\bf Cam Ward} &
+        \multicolumn{2}{c|}{\bf Shedeur Sanders} &
+        \multicolumn{2}{c|}{\bf Jaxson Dart} &
+        \multicolumn{2}{c}{\bf Dillon Gabriel}\\
+      Comp & Score & Comp & Score & Comp & Score & Comp & Score\\
+      \hline
+      \input{../../tables/side_by_side_similarity.tex}
+    \end{tabular}
+  }
+  \caption{\textit{Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft.
+The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}}
+  \label{tab:side-by-side-similarity}
 \end{table}
 
-Figure \ref{fig:prospect-plots} shows the weights of the QBRs used to calculate the predictions, with a green line representing the value of the prediction.
-There is also a graph that shows the similarity scores vs.
-QBR value.
-These plots were produced for each of the four 2025 prospects.
-The scatterplots show the training player similarity score against their QBR for each of the prospects.
-We can see that there is a large concentration of training players with a value of 0 for their QBR value.
-This results from the training dataset containing 2,114 players, with only 300 of those players recording NFL stats (meaning the rest have a value of 0 for the mean QBR).
-We can see that besides the high concentration of training players at zero, Cam Ward has the highest concentration of players around 55-65, which is why his prediction is the highest.
-On the opposite end of the spectrum, Jaxson Dart has the highest concentration of training players at around zero and the lowest concentration at relatively high QBR values.
-This explains why his prediction was by far the lowest of the four prospects we looked at.
+It is worth noting that both Ward and Sanders started their college careers at teams that are not in the NCAA Division I FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State.
+As a result, their college statistics used in the model do not include their statistics from these schools.
+We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
+Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect.
+However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft.
+Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration.
+He also performed well in his interviews with teams, which is not a quantifiable measure.
+
+Figure \ref{fig:prospect-plots} provides a full breakdown of the random forest predictions based on the each prospect's similarity scores for all of the historical prospects in the training data.
+Each prospect's prediction is a weighted average of past prospect's outcomes, and the weighted histograms in Figure \ref{fig:prospect-plots} visualize that weighted mean.
+We observe that all four 2025 prospects have a large concentration of weight on training players with 0 QBR, a reminder that even top prospects have a danger of becoming busts.
+Ward has the highest concentration of comps between 55 and 65 QBR, which is why his prediction is the highest.
+On the opposite end of the spectrum, Dart has the highest concentration of training players near 0 QBR, explaining why his prediction was by far the lowest of the four prospects.
 
@@
     \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
-    \caption{\textit{Top: Histogram of training players’ QBR values weighted by the proportion of shared nodes with the prospect (similarity).
-The green vertical line represents prospect’s random forest predicted QBR based on the similarity-weighted average.
-The red density curve provides a smoothed visualization of the distribution.
-    Bottom: Scatter plot showing similarity scores for training players plotted against their actual QBR values.
-Higher points indicate greater similarity to the prospect, highlighting which historical players he shared the nodes with in the random forest model.}}
+    \caption{\textit{Top: Histogram of training players' QBR values weighted by similarity to the reference prospect.
+The vertical green line annotates each prospect's predicted QBR, which matches the mean of the histograph distribution.
+Bottom: Scatter plot showing individual similarity scores and QBR for historical prospects.}}
     \label{fig:prospect-plots}
@@
 
-We have demonstrated a methodology that produces similarity scores between draft prospects and current/past NFL players based on college statistics by using the nearest-neighbors properties of the random forest algorithm.
-Comparisons between draft prospects and current players are important because they allow the analytical models to be presented in a more digestible way to fans and front offices.
-They allow front offices to make more informed decisions on draft day.
-Additionally, this method provides a way to interpret and understand the random forest algorithm.
+We have demonstrated a methodology for producing similarity scores to help interpret the predictions of random forests.
+These similarity scores have the nice property that the random forest predictions are the weighted averages of outcomes in the training set, weighted according to these similarity scores.
+In addition to interpreting the random forest predictions, the methodology is helpful for identifying similar players based on the random forest as an adaptive $k$-NN algorithm, meaning that the neighborhoods are chosen in a meaningful way for predicting outcomes, which is not the case for a standard $k$-NN algorithm based on Euclidean distance.
+Our methodology is not limited to quarterbacks, to football, or even to sports, although the interpretation of predictions as a weighted average of historical comps fits well with the tradition of using comps in traditional scouting for sports.
 
-This methodology is not specific to quarterbacks or even professional football.
-There are further applications for other position groups, as well as other sports.
-Similarity scores can be used when scouting running backs, wide receivers, defensive backs, linebackers, or virtually any position with well documented college statistics.
-Beyond football, this methodology can be adapted for basketball, baseball, or soccer and can even be used when considering bringing up players from the minor leagues.
+In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in QBR, which highlights some limitations of statistical forecasting when scouting draft prospects.
+The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance.
+However, these two things also highlight the importance of non-quantifiable factors---such as attitude, coachability, off-the-field issues, adaptability, and leadership.
 
-This model's performance means that close to one third of the variance in QBR is explained by these variables, which highlights some limitations of statistical forecasting when scouting draft prospects.
-The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance.
-However, these two things also highlight the importance of non-quantifiable factors – such as attitude, coachability, off-the-field issues, adaptability, and leadership.
+[INSERT PARAGRAPH ABOUT SOFTWARE]
+
+[INSERT PARAGRAPH ABOUT FUTUR WORK EXTENDING TO OTHER TREE-BASED METHODS]
 

### 2025-06-04 scripts/generate_results.R
replace last conference with strength of schedule

@@ -8,6 +8,8 @@ library(xtable)
 
 cfb_data <- read.csv("data/cfb_passing_stats_1984_2024.csv") 
 
+sos_data <- read.csv("data/cfb_schedule_strength_2000_2024.csv")
+
 cfb_rush <- read.csv("data/cfb_rushing_stats_2000_2024.csv") %>%
   mutate(Player = str_replace(Player, "\\*", "")) %>%
   select(Player, Season, Team, rush_att = Att, rush_yds = Yds, rush_td = TD)
@@ -29,6 +31,7 @@ cfb_data <- cfb_data %>%
            grepl("H-9", Awards) ~ 9
          )) %>% 
   left_join(cfb_rush, by = c("Player", "Season", "Team")) %>%
+  left_join(sos_data, by = c("Team" = "School", "Season")) %>%
   group_by(Player) %>%
   summarise(
     c_career_tot_games = sum(G),
@@ -51,7 +54,8 @@ cfb_data <- cfb_data %>%
     last_conference = last(Conf),
     AA = sum(all_american),
     last_H_vote = last(heisman_voting),
-    won_H = sum(won_heisman)
+    won_H = sum(won_heisman),
+    sos = last(SOS)
   ) 
 
 cfb_career_data <- cfb_data %>%
@@ -72,7 +76,7 @@ QBR_passing_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
 QBR_passing_data_combined <- cfb_career_data %>%
   left_join(QBR_passing_data, by = "Player") %>%
   mutate_all(~ replace(., is.na(.), 0)) %>%
-  filter(last_season > 2000, Att == 0 | !(Att < 10))
+  filter(last_season > 2000, Att == 0 | !(Att < 10)) 
 
 set.seed(123)  # For reproducibility
 
@@ -82,8 +86,8 @@ test_data <- QBR_passing_data_combined[-train_indices, ]
 
 # Model tuning
 task_data <- train_data %>%
-  select(c_career_games, c_career_cmp, c_career_att, c_career_yds, c_career_td, c_career_int, last_games, last_passer_rating, AA, last_H_vote, won_H, last_conference, mean_QBR, yds_per_att, final_yds_per_att, c_rush_att, c_rush_yds, c_rush_td)
-task_data$last_conference <- as.factor(task_data$last_conference)
[…]

### 2025-06-04 articles/arxiv/paper.tex
replace instances of conference with strength of schedule

@@
 Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
-The data for each player-season include team, conference, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
+The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
 
@@
 The response variable was average QBR per season (zero for players who never appeared in the NFL).
-The college statistics used as predictor variables were games played per season, completed passes per season, attempted passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season conference, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
+The college statistics used as predictor variables were games played per season, completed passes per season, attempted passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
 

### 2025-06-06 scripts/generate_results.R
update tuned parameters

@@ -107,7 +107,7 @@ rf_model <- ranger(mean_QBR ~ c_career_games + c_career_cmp + c_career_att + c_c
                    + final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
                    data = train_data,  
                    num.trees = 500, importance = "impurity",
-                   keep.inbag = TRUE, min.node.size = 84, mtry = 10)
+                   keep.inbag = TRUE, min.node.size = 67, mtry = 5)
 
 test_terminal_nodes <- predict(rf_model, data = test_data, type = "terminalNodes")$predictions
 
@@ -185,7 +185,7 @@ full_rf_model <- ranger(mean_QBR ~ c_career_games + c_career_cmp + c_career_att 
                         + final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
                         data = QBR_passing_data_combined,  
                         num.trees = 500, importance = "impurity",
-                        keep.inbag = TRUE, min.node.size = 84, mtry = 10)
+                        keep.inbag = TRUE, min.node.size = 67, mtry = 5)
 
 cfb_career_data_current <- cfb_data %>%
   filter(last_season == 2024) %>%

### 2025-06-06 articles/arxiv/paper.tex
update RMSE and % variance explained

@@
 To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
-After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be X and the optimal minimum node size to be XX.
+After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 5 and the optimal minimum node size to be 67.
 
@@
 Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
-The model's test RMSE was 9.89 units of QBR, corresponding to explaining 35.1\% of the variance of QBR in the test set.
+The model's test RMSE was 10.07 units of QBR, corresponding to explaining 32.9\% of the variance of QBR in the test set.
 Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success.

### 2025-06-22 scripts/generate_results.R
Update generate_results.R

@@ -8,6 +8,8 @@ library(xtable)
 
 cfb_data <- read.csv("data/cfb_passing_stats_1984_2024.csv") 
 
+sos_data <- read.csv("data/cfb_schedule_strength_2000_2024.csv")
+
 cfb_rush <- read.csv("data/cfb_rushing_stats_2000_2024.csv") %>%
   mutate(Player = str_replace(Player, "\\*", "")) %>%
   select(Player, Season, Team, rush_att = Att, rush_yds = Yds, rush_td = TD)
@@ -29,6 +31,7 @@ cfb_data <- cfb_data %>%
            grepl("H-9", Awards) ~ 9
          )) %>% 
   left_join(cfb_rush, by = c("Player", "Season", "Team")) %>%
+  left_join(sos_data, by = c("Team" = "School", "Season")) %>%
   group_by(Player) %>%
   summarise(
     c_career_tot_games = sum(G),
@@ -51,7 +54,8 @@ cfb_data <- cfb_data %>%
     last_conference = last(Conf),
     AA = sum(all_american),
     last_H_vote = last(heisman_voting),
-    won_H = sum(won_heisman)
+    won_H = sum(won_heisman),
+    sos = last(SOS)
   ) 
 
 cfb_career_data <- cfb_data %>%
@@ -59,20 +63,44 @@ cfb_career_data <- cfb_data %>%
          c_career_tot_games > 6,
          c_career_att > 5)
 
-QBR_passing_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
+set.seed(123)
+
+nfl_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
   filter(Pos == "QB") %>%
+  mutate(QBR = ifelse(is.na(QBR), 0, QBR),
+         Att = ifelse(is.na(Att), 0, Att))
+
+QBR2_passing_data <- nfl_data %>%
+  arrange(Player, Season) %>%
   group_by(Player) %>%
-  summarise(mean_QBR = mean(QBR),
-            Att = sum(Att),
-            seasons = n()) %>%
-  mutate(
-    mean_QBR = ifelse(is.na(mean_QBR), 0, mean_QBR)
-  )
+  mutate(QBR2 = lead(QBR),
+         QBR2 = ifelse(is.na(QBR2), 0, QBR2)) %>%
+  filter(Season != 2024) 
+
+QBR2_model <- nls(
[…]

### 2025-06-22 scripts/generate_results.R
Update generate_results.R

@@ -63,20 +63,44 @@ cfb_career_data <- cfb_data %>%
          c_career_tot_games > 6,
          c_career_att > 5)
 
-QBR_passing_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
+set.seed(123)
+
+nfl_data <- read.csv("data/QBR_stats_2006_2024.csv") %>%
   filter(Pos == "QB") %>%
+  mutate(QBR = ifelse(is.na(QBR), 0, QBR),
+         Att = ifelse(is.na(Att), 0, Att))
+
+QBR2_passing_data <- nfl_data %>%
+  arrange(Player, Season) %>%
   group_by(Player) %>%
-  summarise(mean_QBR = mean(QBR),
-            Att = sum(Att),
-            seasons = n()) %>%
-  mutate(
-    mean_QBR = ifelse(is.na(mean_QBR), 0, mean_QBR)
-  )
+  mutate(QBR2 = lead(QBR),
+         QBR2 = ifelse(is.na(QBR2), 0, QBR2)) %>%
+  filter(Season != 2024) 
+
+QBR2_model <- nls(
+  formula = QBR2 ~ (Att * QBR) / (Att + c),
+  data = QBR2_passing_data,
+  start = list(c = 1)  # Starting guess for c
+)
+
+summary(QBR2_model)
+
+c <- coef(QBR2_model)[1]
+
+QBR_passing_data <- nfl_data %>%
+  group_by(Player) %>%
+  summarise(
+    Career_Att = sum(Att),
+    mean_QBR = mean(QBR),
+    seasons = n(),
+    Avg_Att = mean(Att)
+  ) %>%
+  mutate(reg_QBR = ((Avg_Att * mean_QBR) / (Avg_Att + c)))
 
 QBR_passing_data_combined <- cfb_career_data %>%
   left_join(QBR_passing_data, by = "Player") %>%
   mutate_all(~ replace(., is.na(.), 0)) %>%
-  filter(last_season > 2000, Att == 0 | !(Att < 10)) 
+  filter(last_season > 2000) 
 
 set.seed(123)  # For reproducibility
 
@@ -86,9 +110,9 @@ test_data <- QBR_passing_data_combined[-train_indices, ]
 
 # Model tuning
 task_data <- train_data %>%
[…]

### 2025-06-24 articles/arxiv/paper.tex
add paragraph explaining adjusted QBR

--- 2025-06-04 articles/arxiv/paper.tex
@@
 
+Since it is possible for a quarterback with very few pass attempts to have a high QBR, we adjusted the average QBR by regressing it toward zero based on attempts per season.
+We used attempts per season rather than career attempts to prevent discrimination against players who have not yet played many seasons.
+We used the formula $QBR_{adj}=/frac{avg_att*\bar{QBR}}{avg_att+c}$ for regression, where c is a constant.
+We used a nonlinear least squares model with a formula of $QBR2~/frac{(Att * QBR)}{(Att + c)}$, where Att is that season's total attempts, QBR is that season's QBR, and QBR2 is the next season's QBR, to find c, resulting in a value of 22.50.
+
 Average QBR per season is the response variable describing the success of quarterback prospects.
@@
 To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
-After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be X and the optimal minimum node size to be XX.
+After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 3 and the optimal minimum node size to be 21.
 
@@
 Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
-The model's test RMSE was 9.89 units of QBR, corresponding to explaining 35.1\% of the variance of QBR in the test set.
+The model's test RMSE was 8.61 units of QBR, corresponding to explaining 33.3\% of the variance of QBR in the test set.
 Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success.
@@
     \includegraphics[width=0.75\linewidth]{../../figures/variable_importance.pdf}
-    \caption{\textit{Variable importance plot for random forest trained to predict NFL QBR for college quarterbacks.}}
+    \caption{\textit{Variable importance plot for random forest trained to predict adjusted NFL QBR for college quarterbacks.}}
     \label{fig:importance}
@@
 
-Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted QBR against career passing yards per season and final-season passer rating.
+Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted adjusted QBR against career passing yards per season and final-season passer rating.
 These are two of the most important predictors in the model, and they have a relatively low correlation.
 The extreme outlier by passing yards per season is Bailey Zappe, who played only one FBS season at Western Kentucky in Conference USA and was selected by the New England Patriots in the fourth round of the 2022 NFL Draft.
-The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed a QBR of 40.
+The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed an adjusted QBR of 40.
 This speaks to the large number of college quarterbacks who achieve zero QBR in the NFL and the difficulty of distinguishing between prospects based on college performance statistics alone.
@@
     \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
-    \caption{\textit{Predicted QBR plotted against average passing yards per season and final-season passer rating.}}
+    \caption{\textit{Predicted adjusted QBR plotted against average passing yards per season and final-season passer rating.}}
     \label{fig:3d-plot}
@@
 Figure \ref{fig:predicted-vs-actuals} shows actual vs.
-predicted QBR for all quarterbacks in the dataset.
+predicted adjusted QBR for all quarterbacks in the dataset.
 Recall that 91\% of actual QBRs are zero, which the visualization does not make clear.
-The errors are very right-skewed, and this illustrates why no college quarterback exceed a QBR prediction of 50.
+The errors are very right-skewed, and this illustrates why no college quarterback exceed an adjusted QBR prediction of 50.
 
@@
   \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
-  \caption{\textit{Predicted (from random forest) and actual QBR for all quarterbacks in the dataset.
+  \caption{\textit{Predicted (from random forest) and actual adjusted QBR for all quarterbacks in the dataset.
 The vats majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}}
@@
 
-Table \ref{tab:top-ten} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
-Cam Ward is predicted to have the highest QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft.
-Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest QBRs, respectively.
+Table \ref{tab:top-ten} shows the adjusted QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
+Cam Ward is predicted to have the highest adjusted QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft.
+Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest adjusted QBRs, respectively.
 While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025.
@@
   \end{tabular}
-  \caption{\textit{Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class.
+  \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class.
 This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
@@
 As a result, their college statistics used in the model do not include their statistics from these schools.
-We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
+We observe that Ward has the highest adjusted QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
 Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect.
@@
     \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
-    \caption{\textit{Top: Histogram of training players' QBR values weighted by similarity to the reference prospect.
-The vertical green line annotates each prospect's predicted QBR, which matches the mean of the histograph distribution.
-Bottom: Scatter plot showing individual similarity scores and QBR for historical prospects.}}
+    \caption{\textit{Top: Histogram of training players' adjusted QBR values weighted by similarity to the reference prospect.
+The vertical green line annotates each prospect's predicted adjusted QBR, which matches the mean of the histograph distribution.
+Bottom: Scatter plot showing individual similarity scores and adjusted QBR for historical prospects.}}
     \label{fig:prospect-plots}
@@
 
-In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in QBR, which highlights some limitations of statistical forecasting when scouting draft prospects.
+In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in adjusted QBR, which highlights some limitations of statistical forecasting when scouting draft prospects.
 The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance.

### 2025-07-06 articles/arxiv/paper.tex
Fix typo
add abstract

@@
 \begin{abstract}
-  The abstract goes here.
+  Random forests are a powerful yet opaque machine learning predictive model.
+This poses challenges for interpretability, particularly in environments where decisions can be very high-stakes, like professional sports.
+In this paper, we present a method for interpreting random forest predictions in a sports setting through the lens of player comps.
+Player comps are a common practice in player scouting, where prospects are evaluated based on their perceived similarity to past players.
+We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors.
+A player’s similarity scores can be viewed as quantifiable player comps.
+We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN’s Total Quarterback Rating (QBR) per season, which has been regressed to zero based on attempts per season as the outcome variable.
+We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.
 \end{abstract}
@@
 While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025.
-Jackson Dart's draft stock increased substantially after the Scouting Combine and his pro day.
+Jaxson Dart's draft stock increased substantially after the Scouting Combine and his pro day.
 

### 2025-07-09 articles/arxiv/paper.tex
refine abstract

@@
 We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors.
-A player’s similarity scores can be viewed as quantifiable player comps.
-We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN’s Total Quarterback Rating (QBR) per season, which has been regressed to zero based on attempts per season as the outcome variable.
+A player’s similarity scores can be viewed as quantifiable player comps, and the random forest's prediction can be recovered by taking a weighted average of the outcome variable among those comps, with weights determined by the similarity scores.
+We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN’s Total Quarterback Rating (QBR) per season.
 We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.

### 2025-07-16 scripts/generate_results.R
create R package

@@ -203,13 +203,17 @@ new_plot_data <- QBR_passing_data_combined %>%
   dev.off()
 }
 
-full_rf_model <- ranger(reg_QBR ~ c_career_games + c_career_cmp + c_career_att + c_career_yds 
-                        + c_career_td + c_career_int + last_games + last_passer_rating + AA 
-                        + last_H_vote + won_H + sos + yds_per_att
-                        + final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
-                        data = QBR_passing_data_combined,  
-                        num.trees = 500, importance = "impurity",
-                        keep.inbag = TRUE, min.node.size = 50, mtry = 5)
+full_rf_model <- ranger(
+  formula = reg_QBR ~ c_career_games + c_career_cmp + c_career_att + c_career_yds + c_career_td +
+    c_career_int + last_games + last_passer_rating + AA + last_H_vote + won_H + sos + yds_per_att +
+    final_yds_per_att + c_rush_att + c_rush_yds + c_rush_td,
+  data = QBR_passing_data_combined,
+  num.trees = 500,
+  importance = "impurity",
+  keep.inbag = TRUE,
+  min.node.size = 50,
+  mtry = 5
+)
 
 cfb_career_data_current <- cfb_data %>%
   filter(last_season == 2024) %>%
@@ -226,49 +230,11 @@ cfb_career_data_current %>%
   mutate(predictions = sprintf("%.1f", predictions)) %>%
   sputil::write_latex_table(file = "tables/top_ten.tex")
 
-
-test_names <- cfb_career_data_current$Player
-test_terminal_nodes <- predict(full_rf_model, data = cfb_career_data_current, type = "terminalNodes")$predictions
-rownames(test_terminal_nodes) <- test_names
-train_terminal_nodes <- predict(full_rf_model, data = QBR_passing_data_combined, type = "terminalNodes")$predictions
-rownames(train_terminal_nodes) <- QBR_passing_data_combined$Player
-inbag_counts <- full_rf_model$inbag.counts
-inbag_counts <- do.call(cbind, inbag_counts)
-
-M <- ncol(train_terminal_nodes)  # Number of trees
-n_test <- nrow(cfb_career_data_current)  # Number of test samples
-n_train <- nrow(QBR_passing_data_combined)  # Number of training samples
-
-# Initialize similarity matrix (rows = test players, cols = training players)
[…]

### 2025-07-16 R/extract_similarity.R
Create R package (#1)

@@ -1,0 +1,108 @@
+#' Extract Random Forest Similarity Scores
+#'
+#' Calculates similarity scores between new observations and reference data points based on a
+#' fitted random forest model. The similarity reflects the frequency with which each new data point
+#' shares terminal nodes with each reference data point across the forest, weighted appropriately.
+#' This corresponds to the adaptive nearest neighbors interpretation of random forests.
+#'
+#' Supports random forest models fitted with either the \pkg{ranger} or \pkg{randomForest} package.
+#'
+#' @param object A fitted random forest model. Must be an object of class \code{"ranger"} (from the
+#' \pkg{ranger} package) or \code{"randomForest"} (from the \pkg{randomForest} package).
+#' @param newdata A data frame or matrix of new observations to compare against \code{refdata}.
+#' @param refdata A data frame or matrix containing the reference data points (e.g., training data
+#' or held-out data).
+#' @param match_training Logical. If \code{TRUE}, uses in-bag sampling frequencies to weight the
+#' similarity scores based on how often each reference point appeared in each bootstrap sample.
+#' Must be \code{TRUE} if \code{refdata} was used to train the model and you want the similarity to
+#' reflect the adaptive nearest neighbor weighting. For this to work, \code{object} must have kept
+#' track of the training inbag counts, i.e. \code{keep.inbag = TRUE}, which is not default behavior.
+#'
+#' @return A numeric matrix of dimension \code{nrow(newdata)} by \code{nrow(refdata)}, where each
+#' entry represents the similarity score between a new point and a reference point. The rows sum
+#' (approximately) to 1.
+#'
+#' @details
+#' For each tree in the forest, new and reference points are considered similar if they fall into
+#' the same terminal node. Similarity scores are calculated as the normalized frequency with which
+#' pairs share terminal nodes, adjusted by the in-bag counts of reference points when
+#' \code{match_training = TRUE}. This weighting ensures that predictions made using these
+#' similarity scores match the forest's behavior when trained using bootstrap samples.
+#'
[…]

### 2025-07-18 articles/arxiv/paper.tex
use biblatex instead of natbib

--- 2025-07-06 articles/arxiv/paper.tex
@@
 
-\usepackage{amsmath, amsfonts, array, authblk, float, graphicx, hyperref, natbib, subcaption}
+\usepackage[alldates=year, block=space, style=apa]{biblatex}
+\setlength\bibitemsep{0.5\baselineskip}
+\addbibresource{../random-forest-comps.bib}
+
 \usepackage[margin=1in]{geometry}
+\linespread{1.2}
 
-\bibliographystyle{../apa-good}
-\linespread{1.2}
+\usepackage[colorlinks, allcolors=blue]{hyperref}
+
+\usepackage{amsmath, amsfonts, array, authblk, float, graphicx, subcaption}
 
@@
 We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors.
-A player’s similarity scores can be viewed as quantifiable player comps.
-We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN’s Total Quarterback Rating (QBR) per season, which has been regressed to zero based on attempts per season as the outcome variable.
+A player's similarity scores can be viewed as quantifiable player comps, and the random forest's prediction can be recovered by taking a weighted average of the outcome variable among those comps, with weights determined by the similarity scores.
+We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating (QBR) per season.
 We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.
@@
 Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks).
-Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.
+Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data (\cite{breiman_random_2001}), which makes them suitable for many real-world applications.
 
 Like other so-called ``black box'' machine learning algorithms, one drawback of the random forest is that the reasons for an individual's prediction can be difficult to interpret.
-Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction \citep{ribeiro_why_2016}.
+Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction (\cite{ribeiro_why_2016}).
 This is especially important in sports contexts because executives need to consider non-quantifiable information in their decision-making.
@@
 Perhaps one of the most important decisions that sports executives make is selecting quarterbacks in the National Football League (NFL) draft.
-Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position \citep{hughes_positional_2015}.
+Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position (\cite{hughes_positional_2015}).
 Due to the nature of the position, trading for or signing a solid quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
@@
 In 2025, the slotted contract for No.
-1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus \citep{badenhausen_nfl_2025}.
+1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus (\cite{badenhausen_nfl_2025}).
 
 In traditional scouting for sports, draft prospect discussions and predictions are often based on comparing a current prospect with similar draft prospects in the past.
-Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success \citep{trapasso_nfl_2025}.
+Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success (\cite{trapasso_nfl_2025}).
 Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers.
-These historical reference players are known colloquially as ``comps'' \citep{jones_nfl_2025}, short for comparables.
+These historical reference players are known colloquially as ``comps'' (\cite{jones_nfl_2025}), short for comparables.
 The simplest example of this line of reasoning is, ``This prospect reminds me most of Player X, and Player X had a successful career, so I think this player will have a successful career,'' effectively a human implementation of the $k$-nearest neighbor ($k$-NN) algorithm, with $k = 1$.
 
-Random forests can be thought of as an adaptive $k$-NN algorithm \citep{lin_random_2006}, with the nearness of neighbors adaptively determined to minimize prediction error.
+Random forests can be thought of as an adaptive $k$-NN algorithm (\cite{lin_random_2006}), with the nearness of neighbors adaptively determined to minimize prediction error.
 Our work is moivated by this connection between random forests and the traditional lens of player comps.
@@
 Previous research has been done to attempt to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
-\citet{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
-\citet{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models.
-In a sports context, \citet{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.
+\textcite{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
+\textcite{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models.
+In a sports context, \textcite{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.
 
 Draft models are fairly common in professional sports, as evaluating prospect talent can be key to success.
-One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \citet{craig_predicting_2021}.
+One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \textcite{craig_predicting_2021}.
 They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season metrics, scout grades, and player height.
 Their predictions are easy to interpret directly from the logistic regression model.
-\citet{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
+\textcite{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
 They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting a quarterback.
 Again, the logistic regression model results in predictions that are easily interpretable.
-\citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
+\textcite{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
 The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model.
-\citet{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics.
+\textcite{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics.
 They use a combination of large language models and ensemble machine learning, which are both opaque in the way that they generate predictions, leading to uninterpretable predictions/reports.
@@
 
-All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}.
+All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}).
 Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
@@
 
-In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) \citep{burke_how_2016} from each season between 2006 and 2024 (inclusive).
+In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) (\cite{burke_how_2016}) from each season between 2006 and 2024 (inclusive).
 Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play.
 It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time.
-The range of QBR is from 0 to 100, and \citet{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
+The range of QBR is from 0 to 100, and \textcite{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
 For each quarterback, we calculate their average QBR per season, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero.
@@
 
-Random forests are a machine learning method typically used for classification or regression \citep{breiman_random_2001}.
+Random forests are a machine learning method typically used for classification or regression (\cite{breiman_random_2001}).
 The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision within each bootstrap sample, randomly choosing at each node a subset of the features to consider for the split.
@@
 
-We trained the random forest model using the R package ranger \citep{wright_ranger_2015}, which is a fast implementation of the random forest algorithm.
+We trained the random forest model using the R package ranger (\cite{wright_ranger_2015}), which is a fast implementation of the random forest algorithm.
 The response variable was average QBR per season (zero for players who never appeared in the NFL).
@@
 We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement.
-To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
+To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2018}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
 After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 3 and the optimal minimum node size to be 21.
@@
 
-\citet{lin_random_2006} previously established the relationship between $k$-NN and the random forest algorithm.
+\textcite{lin_random_2006} previously established the relationship between $k$-NN and the random forest algorithm.
 In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data.
@@
 Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft.
-However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance \citep{mckenna_what_nodate}.
+However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_nodate}).
 In the end, Dart was the other quarterback taken in the first round of the draft.
@@
 
-\bibliography{../random-forest-comps}
+\printbibliography
 

### 2025-07-18 articles/arxiv/paper.tex
draft software section

@@
 
+The primary contribution of the present work is less about the specific application to quarterback prospects in football and more about introducing the concept of extracting similarity scores from random forests to interpret the predictions.
+To this end, we have published the R package treecomp, which provides a simple function interface for extracting similarity scores from fitted random forest objects.
+Currently, the package handles fitted random forests from the ranger and randomForest packages.
+The \texttt{extract\_similarity} function has syntax similar to a \texttt{predict} method and requires as arguments only the fitted random forest object, the new data (for whom comps are to be found) and the reference data (the candidate comps).
+If the reference data are the same as the training data, then the predictions of the random forest can be recovered as the weighted average of the training outcomes, weighted by similarity score.
+However, it is also possible to use a different dataset as the reference data; in this case, the similarity is still interpretable as corresponding to the adaptive nearness learned by the random forest.
+This adaptive nearness is more informative than Euclidean distance because it reflects nearness in ways that are important for predicting outcomes.
+
+The treecomp R package is available at \url{https://github.com/elisabethmill/treecomp}.
+
 \section{Discussion}

### 2025-07-21 scripts/generate_results.R
fix filter issues

@@ -100,7 +100,7 @@ QBR_passing_data <- nfl_data %>%
 QBR_passing_data_combined <- cfb_career_data %>%
   left_join(QBR_passing_data, by = "Player") %>%
   mutate_all(~ replace(., is.na(.), 0)) %>%
-  filter(last_season > 2000) 
+  filter(last_season > 2000, Avg_Att == 0 | !(Avg_Att < 10)) 
 
 set.seed(123)  # For reproducibility
 

### 2025-07-21 scripts/generate_results.R
update results script with new variable names

@@ -10,15 +10,15 @@ set.seed(123)
 
 # Determine mean regression for QBR ----
 
-QBR2_passing_data <- treecomp::NFL_QBR_by_year %>%
-  arrange(Player, Season) %>%
-  group_by(Player) %>%
-  mutate(QBR2 = lead(QBR),
-         QBR2 = ifelse(is.na(QBR2), 0, QBR2)) %>%
-  filter(Season != 2024)
+QBR2_passing_data <- treecomp::nfl_qbr_by_year %>%
+  arrange(player_name, year) %>%
+  group_by(player_name) %>%
+  mutate(qbr_2 = lead(qbr),
+         qbr_2 = ifelse(is.na(qbr_2), 0, qbr_2)) %>%
+  filter(year != 2024)
 
 QBR2_model <- nls(
-  formula = QBR2 ~ (Att * QBR) / (Att + c),
+  formula = qbr_2 ~ (att * qbr) / (att + c),
   data = QBR2_passing_data,
   start = list(c = 1)  # Starting guess for c
 )
@@ -32,18 +32,18 @@ c <- coef(QBR2_model)[1]
 
 data <- treecomp::quarterback %>%
   mutate(
-    reg_QBR = Avg_Att / (Avg_Att + c) * mean_QBR,
+    reg_qbr = nfl_att_per_year / (nfl_att_per_year + c) * nfl_qbr_per_year,
     # NA for Heisman voting means you didn't finish in the top 10
-    last_H_vote = ifelse(is.na(last_H_vote), 11, last_H_vote)
+    ncaa_heisman_last = ifelse(is.na(ncaa_heisman_last), 11, ncaa_heisman_last)
   )
 
 past_data <- data %>%
   filter(
-    last_season < 2024,
-    c_career_tot_games > 6,
-    c_career_att > 5,
-    # Exclude players with very few attempts per season because their QBR will be unstable
-    is.na(Avg_Att) | (Avg_Att == 0) | Avg_Att >= 10
+    ncaa_year_last < 2024,
+    ncaa_games_career > 6,
+    ncaa_att_per_year > 5,
+    # Exclude players with very few NFL attempts per season because their QBR will be unstable
+    is.na(nfl_att_per_year) | (nfl_att_per_year == 0) | nfl_att_per_year >= 10
   ) %>%
   # Remaining NAs are due to not playing in the NFL
   mutate_all(~ replace(., is.na(.), 0))
[…]

### 2025-07-21 R/data.R
update data documentation

@@ -6,33 +6,33 @@
 #' @format ## `quarterback`
 #' A data frame with 5,701 rows and 27 columns:
 #' \describe{
-#'   \item{Player}{Player name}
-#'   \item{ncaa_seasons}{Number of seasons played in college career}
-#'   \item{c_career_tot_games}{Total number of games in college career}
-#'   \item{c_career_games}{Games per year in college career}
-#'   \item{c_career_att}{Passing attempts per year in college career}
-#'   \item{c_career_cmp}{Passing completions per year in college career}
-#'   \item{c_career_yds}{Passing yards completions per year in college career}
-#'   \item{c_career_td}{Passing touchdowns per year in college career}
-#'   \item{c_career_int}{Passing interceptions per year in college career}
-#'   \item{c_rush_att}{Rushing attempts per year in college career}
-#'   \item{c_rush_yds}{Rusing yards per year in college career}
-#'   \item{c_rush_td}{Rushing touchdowns per year in college career}
-#'   \item{yds_per_att}{Passing yards per attempt in college career}
-#'   \item{final_yds_per_att}{Passing yards per attempt in final college season}
-#'   \item{last_season}{Year of final college season}
-#'   \item{last_conference}{Conference for final college season}
-#'   \item{last_college}{Team for final college season}
-#'   \item{last_sos}{Strength of schedule for final college season}
-#'   \item{last_games}{Games played in final college season}
-#'   \item{last_passer_rating}{Passer rating in final college season}
-#'   \item{AA}{Numer of All-America recognitions in college career}
-#'   \item{won_H}{Number of Heisman awards in college career}
-#'   \item{last_H_vote}{Heisman voting finish in final college season}
-#'   \item{nfl_seasons}{Number of seasons played in NFL career}
-#'   \item{Career_Att}{Total passing attempts in NFL career}
-#'   \item{Avg_Att}{Passing attempts per year in NFL career}
-#'   \item{mean_QBR}{Average QBR (across years) in NFL career}
+#'   \item{player_name}{Player name}
+#'   \item{ncaa_years_career}{Number of seasons played in college career}
+#'   \item{ncaa_games_career}{Total number of games in college career}
+#'   \item{ncaa_yds_per_att_career}{Passing yards per attempt in college career}
[…]

### 2025-07-21 scripts/download_data.R
Choose clear variable names for broad audience (#4)
* make variable names clearer

* update package data

* update results script with new variable names

* update data documentation

@@ -127,7 +127,7 @@ if (verbose) {
   logger::log_info("Downloading NFL QBR data (2006-2024)")
 }
 
-NFL_QBR_by_year <- tibble::tibble()
+nfl_qbr_by_year <- tibble::tibble()
 
 for (year in 2006:2024) {
   if (verbose) {
@@ -153,9 +153,9 @@ for (year in 2006:2024) {
     slice(1) %>%
     ungroup() %>%
     mutate(Season = year) %>%
-    select(Player, Season, Att, QBR)
+    select(player_name = Player, year = Season, att = Att, qbr = QBR)
   
-  NFL_QBR_by_year <- bind_rows(NFL_QBR_by_year, year_data)
+  nfl_qbr_by_year <- bind_rows(nfl_qbr_by_year, year_data)
   
   Sys.sleep(runif(1, 8, 10))
 }
@@ -167,13 +167,13 @@ if (verbose) {
 
 # Join all the data together ----
 
-NFL_QBR_career <- NFL_QBR_by_year %>%
-  group_by(Player) %>%
+nfl_qbr_career <- nfl_qbr_by_year %>%
+  group_by(player_name) %>%
   summarize(
-    nfl_seasons = n(),
-    Career_Att = sum(Att),
-    Avg_Att = mean(Att),
-    mean_QBR = mean(QBR)
+    nfl_years = n(),
+    nfl_att_career = sum(att),
+    nfl_att_per_year = mean(att),
+    nfl_qbr_per_year = mean(qbr)
   )
 
 quarterback <- passing_data %>%
@@ -197,36 +197,36 @@ quarterback <- passing_data %>%
       grepl("H-10", Awards) ~ 10
     )
   ) %>%
-  group_by(Player) %>%
+  group_by(player_name = Player) %>%
   summarize(
-    ncaa_seasons = n(),
-    c_career_tot_games = sum(G),
-    c_career_games = mean(G, na.rm = TRUE),
-    c_career_att = mean(Att, na.rm = TRUE),
-    c_career_cmp = mean(Cmp, na.rm = TRUE),
-    c_career_yds = mean(Yds, na.rm = TRUE),
-    c_career_td = mean(TD, na.rm = TRUE),
-    c_career_int = mean(Int, na.rm = TRUE),
-    c_rush_att = mean(rush_att, na.rm = TRUE),
-    c_rush_yds = mean(rush_yds, na.rm = TRUE),
-    c_rush_td = mean(rush_td, na.rm = TRUE),
[…]

### 2025-07-24 scripts/generate_results.R
use the hyperparameters recommended by tuneRanger

@@ -89,8 +89,8 @@ rf_model <- ranger(
   num.trees = 500,
   importance = "impurity",
   keep.inbag = TRUE,
-  min.node.size = 50,
-  mtry = 5
+  min.node.size = tuned_model$recommended.pars$min.node.size,
+  mtry = tuned_model$recommended.pars$mtry
 )
 
 test_terminal_nodes <- predict(rf_model, data = test_data, type = "terminalNodes")$predictions
@@ -177,8 +177,8 @@ full_rf_model <- ranger(
   num.trees = 500,
   importance = "impurity",
   keep.inbag = TRUE,
-  min.node.size = 50,
-  mtry = 5
+  min.node.size = tuned_model$recommended.pars$min.node.size,
+  mtry = tuned_model$recommended.pars$mtry
 )
 
 present_data <- data %>%

### 2025-07-24 scripts/generate_results.R
up the number of trees to 1000

@@ -5,6 +5,8 @@ library(mlr)
 library(ggplot2)
 library(xtable)
 
+num_trees <- 1000
+
 set.seed(123)
 
 
@@ -69,7 +71,7 @@ task <- makeRegrTask(data = task_data, target = "reg_qbr")
 
 tuned_model <- tuneRanger(
   task,
-  num.trees = 500,
+  num.trees = num_trees,
   iters = 100,
   tune.parameters = c("mtry", "min.node.size"),
   parameters = list(replace = TRUE)
@@ -86,7 +88,7 @@ rf_model <- ranger(
     ncaa_sos_last + ncaa_games_last + ncaa_yds_per_att_last + ncaa_passer_rating_last +
     ncaa_all_america + ncaa_heisman + ncaa_heisman_last,
   data = train_data,
-  num.trees = 500,
+  num.trees = num_trees,
   importance = "impurity",
   keep.inbag = TRUE,
   min.node.size = tuned_model$recommended.pars$min.node.size,
@@ -174,7 +176,7 @@ full_rf_model <- ranger(
     ncaa_sos_last + ncaa_games_last + ncaa_yds_per_att_last + ncaa_passer_rating_last +
     ncaa_all_america + ncaa_heisman + ncaa_heisman_last,
   data = past_data,
-  num.trees = 500,
+  num.trees = num_trees,
   importance = "impurity",
   keep.inbag = TRUE,
   min.node.size = tuned_model$recommended.pars$min.node.size,

### 2025-07-24 scripts/generate_results.R
refit the full RF model before producing figures and tables

@@ -95,10 +95,7 @@ rf_model <- ranger(
   mtry = tuned_model$recommended.pars$mtry
 )
 
-test_terminal_nodes <- predict(rf_model, data = test_data, type = "terminalNodes")$predictions
-
 test_data$predictions <- predict(rf_model, data = test_data)$predictions
-
 rmse <- sqrt(mean((test_data$reg_qbr - test_data$predictions)^2))
 total_variance <- var(test_data$reg_qbr)
 residuals <- test_data$reg_qbr - test_data$predictions
@@ -106,9 +103,26 @@ residual_variance <- var(residuals)
 explained_variance <- total_variance - residual_variance
 print((explained_variance / total_variance) * 100)
 
+
+# Re-fit random forest on full data set ----
+
+full_rf_model <- ranger(
+  formula = reg_qbr ~
+    ncaa_yds_per_att_career + ncaa_games_per_year +
+    ncaa_att_per_year + ncaa_cmp_per_year + ncaa_yds_per_year + ncaa_td_per_year + ncaa_int_per_year +
+    ncaa_rush_att_per_year + ncaa_rush_yds_per_year + ncaa_rush_td_per_year +
+    ncaa_sos_last + ncaa_games_last + ncaa_yds_per_att_last + ncaa_passer_rating_last +
+    ncaa_all_america + ncaa_heisman + ncaa_heisman_last,
+  data = past_data,
+  num.trees = num_trees,
+  importance = "impurity",
+  keep.inbag = TRUE,
+  min.node.size = tuned_model$recommended.pars$min.node.size,
+  mtry = tuned_model$recommended.pars$mtry
+)
+
 # Model plots
-importance_scores <- importance(rf_model)
-importance(rf_model)
+importance_scores <- importance(full_rf_model)
 names(importance_scores) <- c("games/season", "completions/season", "attempts/season", "yards/season", "touchdowns/season", "interceptions/season", "final season games", "final season passer rating", "All-American seasons", "final season Heisman voting", "won Heisman Award", "final season strength of schedule", "yards/attempt", "final season yards/attempt", "rushing attempts/season", "rushing yards/season", "rushing touchdowns/season")
 importance_df <- data.frame(
   Variable = names(importance_scores),
@@ -128,7 +142,7 @@ importance_df <- data.frame(
 }
 
 new_plot_data <- past_data %>%
[…]

### 2025-07-24 scripts/generate_results.R
fix variable names for importance plot

@@ -64,7 +64,7 @@ task_data <- train_data %>%
     ncaa_att_per_year, ncaa_cmp_per_year, ncaa_yds_per_year, ncaa_td_per_year, ncaa_int_per_year,
     ncaa_rush_att_per_year, ncaa_rush_yds_per_year, ncaa_rush_td_per_year,
     ncaa_sos_last, ncaa_games_last, ncaa_yds_per_att_last, ncaa_passer_rating_last,
-    ncaa_all_america, ncaa_heisman, ncaa_heisman_last,
+    ncaa_all_america, ncaa_heisman, ncaa_heisman_last
   )
 task_data$ncaa_sos_last <- as.numeric(task_data$ncaa_sos_last)
 task <- makeRegrTask(data = task_data, target = "reg_qbr")
@@ -123,9 +123,27 @@ full_rf_model <- ranger(
 
 # Model plots
 importance_scores <- importance(full_rf_model)
-names(importance_scores) <- c("games/season", "completions/season", "attempts/season", "yards/season", "touchdowns/season", "interceptions/season", "final season games", "final season passer rating", "All-American seasons", "final season Heisman voting", "won Heisman Award", "final season strength of schedule", "yards/attempt", "final season yards/attempt", "rushing attempts/season", "rushing yards/season", "rushing touchdowns/season")
+variable_display <- c(
+  ncaa_yds_per_att_career = "Career Yds/Att",
+  ncaa_games_per_year = "Games/Season",
+  ncaa_att_per_year = "Attempts/Season",
+  ncaa_cmp_per_year = "Completions/Season",
+  ncaa_yds_per_year = "Yards/Season",
+  ncaa_td_per_year = "Touchdowns/Season",
+  ncaa_int_per_year = "Interceptions/Season",
+  ncaa_rush_att_per_year = "Rush Attempts/Season",
+  ncaa_rush_yds_per_year = "Rush Yards/Season",
+  ncaa_rush_td_per_year = "Rush Touchdowns/Season",
+  ncaa_sos_last = "Final Strength of Schedule",
+  ncaa_games_last = "Final Games",
+  ncaa_yds_per_att_last = "Final Yds/Att",
+  ncaa_passer_rating_last = "Final Passer Rating",
+  ncaa_all_america = "All-America Seasons",
+  ncaa_heisman = "Won Heisman Award",
+  ncaa_heisman_last = "Final Heisman Voting"
+)
 importance_df <- data.frame(
[…]

### 2025-07-24 scripts/generate_results.R
Polish random forest model (#5)
* use the hyperparameters recommended by tuneRanger

* up the number of trees to 1000

* refit the full RF model before producing figures and tables

* fix variable names for importance plot

@@ -5,6 +5,8 @@ library(mlr)
 library(ggplot2)
 library(xtable)
 
+num_trees <- 1000
+
 set.seed(123)
 
 
@@ -62,14 +64,14 @@ task_data <- train_data %>%
     ncaa_att_per_year, ncaa_cmp_per_year, ncaa_yds_per_year, ncaa_td_per_year, ncaa_int_per_year,
     ncaa_rush_att_per_year, ncaa_rush_yds_per_year, ncaa_rush_td_per_year,
     ncaa_sos_last, ncaa_games_last, ncaa_yds_per_att_last, ncaa_passer_rating_last,
-    ncaa_all_america, ncaa_heisman, ncaa_heisman_last,
+    ncaa_all_america, ncaa_heisman, ncaa_heisman_last
   )
 task_data$ncaa_sos_last <- as.numeric(task_data$ncaa_sos_last)
 task <- makeRegrTask(data = task_data, target = "reg_qbr")
 
 tuned_model <- tuneRanger(
   task,
-  num.trees = 500,
+  num.trees = num_trees,
   iters = 100,
   tune.parameters = c("mtry", "min.node.size"),
   parameters = list(replace = TRUE)
@@ -86,17 +88,14 @@ rf_model <- ranger(
     ncaa_sos_last + ncaa_games_last + ncaa_yds_per_att_last + ncaa_passer_rating_last +
     ncaa_all_america + ncaa_heisman + ncaa_heisman_last,
   data = train_data,
-  num.trees = 500,
+  num.trees = num_trees,
   importance = "impurity",
   keep.inbag = TRUE,
-  min.node.size = 50,
-  mtry = 5
+  min.node.size = tuned_model$recommended.pars$min.node.size,
+  mtry = tuned_model$recommended.pars$mtry
 )
 
-test_terminal_nodes <- predict(rf_model, data = test_data, type = "terminalNodes")$predictions
-
 test_data$predictions <- predict(rf_model, data = test_data)$predictions
-
 rmse <- sqrt(mean((test_data$reg_qbr - test_data$predictions)^2))
 total_variance <- var(test_data$reg_qbr)
 residuals <- test_data$reg_qbr - test_data$predictions
@@ -104,12 +103,47 @@ residual_variance <- var(residuals)
 explained_variance <- total_variance - residual_variance
 print((explained_variance / total_variance) * 100)
 
+
+# Re-fit random forest on full data set ----
+
+full_rf_model <- ranger(
+  formula = reg_qbr ~
[…]

### 2025-07-24 scripts/download_data.R
store SOS as numeric

@@ -214,7 +214,7 @@ quarterback <- passing_data %>%
     ncaa_year_last = last(Season),
     ncaa_conference_last = last(Conf),
     ncaa_team_last = last(Team),
-    ncaa_sos_last = last(SOS),
+    ncaa_sos_last = as.numeric(last(SOS)),
     ncaa_games_last = last(G),
     ncaa_yds_per_att_last = last(Yds) / last(Att),
     ncaa_passer_rating_last = last(Rate),

### 2025-07-24 scripts/generate_results.R
choose more interesting variables for #D plot

@@ -165,20 +165,19 @@ new_plot_data <- past_data %>%
 {
   sputil::open_device("figures/3d_plot.pdf", height = 5)
   plot <- new_plot_data |>
-    ggplot(aes(x = ncaa_yds_per_year, y = ncaa_passer_rating_last, color = predictions)) +
+    ggplot(aes(x = ncaa_yds_per_year, y = ncaa_sos_last, color = predictions)) +
     geom_point(size = 1) +
     scale_color_viridis_c() +
     labs(
       x = "Passing Yards per Season (College)",
-      y = "Final-Season Passer Rating (College)",
+      y = "Strength of Schedule (College)",
       color = "Predicted QBR"
     ) +
-    coord_cartesian(xlim = c(0, 6000), ylim = c(0, 250)) +
     theme_minimal() +
-    theme(legend.position = "inside", legend.position.inside = c(0.85, 0.25))
+    theme(legend.position = "inside", legend.position.inside = c(0.85, 0.75))
   print(plot)
   dev.off()
-  }
+}
 
 {
   sputil::open_device("figures/predicted_vs_actuals.pdf", height = 5, width = 5)

### 2025-07-24 scripts/generate_results.R
fix latex results table formatting

@@ -209,7 +209,11 @@ present_data %>%
   arrange(-predictions) %>%
   head(10) %>%
   mutate(predictions = sprintf("%.1f", predictions)) %>%
-  sputil::write_latex_table(file = "tables/top_ten.tex")
+  sputil::write_latex_table(
+    file = "tables/top_ten.tex",
+    colnames = c("Player", "Predicted QBR"),
+    align = "l|r"
+  )
 
 
 # Produce similarity figures and tables ----
@@ -438,4 +442,14 @@ ss_top_10_wide <- ss_top_10 %>%
   setNames(c("SS_Player", "SS_Similarity"))
 
 cbind(cw_top_10_wide, ss_top_10_wide, jd_top_10_wide, dg_top_10_wide) %>%
-  sputil::write_latex_table(file = "tables/side_by_side_similarity.tex")
+  sputil::write_latex_table(
+    file = "tables/side_by_side_similarity.tex",
+    colnames = rep(c("Comp", "Score"), times = 4),
+    prefix_rows = "
+      \\multicolumn{2}{c|}{Cam Ward} &
+      \\multicolumn{2}{c|}{Shedeur Sanders} &
+      \\multicolumn{2}{c|}{Jaxson Dart} &
+      \\multicolumn{2}{c}{Dillon Gabriel}
+    ",
+    align = "lr|lr|lr|lr"
+  )

### 2025-07-24 articles/arxiv/paper.tex
fix latex results table formatting
update latest results in paper

@@
 All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}).
-Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
+Our dataset includes 2,099 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
 The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
@@
 
-Since it is possible for a quarterback with very few pass attempts to have a high QBR, we adjusted the average QBR by regressing it toward zero based on attempts per season.
-We used attempts per season rather than career attempts to prevent discrimination against players who have not yet played many seasons.
-We used the formula $QBR_{adj}=/frac{avg_att*\bar{QBR}}{avg_att+c}$ for regression, where c is a constant.
-We used a nonlinear least squares model with a formula of $QBR2~/frac{(Att * QBR)}{(Att + c)}$, where Att is that season's total attempts, QBR is that season's QBR, and QBR2 is the next season's QBR, to find c, resulting in a value of 22.50.
-
 Average QBR per season is the response variable describing the success of quarterback prospects.
@@
 This is a limitation of the data.
+
+Another limitation of average QBR per season is that some players put up very high QBR numbers in very small samples (e.g.
+a QBR of 64.7 on only 26 passing attempts over a full season) and are rated as having very positive outcomes although a more subjective evaluation of their careers would be less positive.
+To mitigate the effect of this phenomenon, we regress each player's career QBR per season toward zero:
+\begin{equation*}
+  \mbox{QBR}_{\mbox{reg}} = \frac{(\mbox{Attempts / Season}) \cdot (\mbox{Average QBR}) + c \cdot 0}{(\mbox{Attempts / Season}) + c}.
+\end{equation*}
+We chose $c = 20.8$ as this value minimizes the error in predicting a player's next-season QBR when used to regrees current-season QBR.
+
+A response variable such as nflWAR (\cite{yurko_nflwar_2019}) or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers.
+The primary purpose of the present work (extracting similarity scores to interpret random forest predictions through the lens of player comps) holds regardless of the response variable used.
 
@@
 To train the random forest model, we first implemented a random 70\%/30\% train/test split.
-We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement.
+We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement.
 To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2018}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
-After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 3 and the optimal minimum node size to be 21.
+After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 5 and the optimal minimum node size to be 73.
 
@@
 Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
-The model's test RMSE was 8.61 units of QBR, corresponding to explaining 33.3\% of the variance of QBR in the test set.
-Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success.
-Note that these three variables are highly correlated with each other---each pairwise correlation is at least 0.94.
-The next most important variables are final-season Heisman voting, average passing attempts per season and final-season passer rating.
+The model's test RMSE was 8.61 units of QBR, corresponding to explaining 43.7\% of the variance of QBR in the test set.
+Final-season Heisman voting, average passing performance (yards, touchdowns, completions) per season, and final-season strength of schedule stand out as the most important predictors of NFL success.
+Note that yards per season, touchdowns per season, and completions per season are highly correlated with each other---each pairwise correlation is at least 0.94.
 
@@
 
-\begin{figure}[H]
+\begin{figure}
     \centering
@@
   \centering
-  \begin{tabular}{l|r}
-    Player & Predicted QBR\\
-    \hline
   \input{../../tables/top_ten.tex}
-  \end{tabular}
   \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class.
-This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
+This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
   \label{tab:top-ten}
@@
   \resizebox{\textwidth}{!}{
-    \begin{tabular}{lr|lr|lr|lr}
-      \multicolumn{2}{c|}{\bf Cam Ward} &
-        \multicolumn{2}{c|}{\bf Shedeur Sanders} &
-        \multicolumn{2}{c|}{\bf Jaxson Dart} &
-        \multicolumn{2}{c}{\bf Dillon Gabriel}\\
-      Comp & Score & Comp & Score & Comp & Score & Comp & Score\\
-      \hline
     \input{../../tables/side_by_side_similarity.tex}
-    \end{tabular}
   }

### 2025-07-24 articles/arxiv/paper.tex
fix mckenna reference with data

@@
 Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft.
-However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_nodate}).
+However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_2025}).
 In the end, Dart was the other quarterback taken in the first round of the draft.

### 2025-07-27 articles/jsa/paper.tex
upload JSA file
update bibliography

--- 2025-07-18 articles/arxiv/paper.tex
@@
-\documentclass{article}
+\documentclass[Review, sageh, times]{sagej}
 
-\usepackage[alldates=year, block=space, style=apa]{biblatex}
-\setlength\bibitemsep{0.5\baselineskip}
+\bibliographystyle{SageH}
 \addbibresource{../random-forest-comps.bib}
-
-\usepackage[margin=1in]{geometry}
-\linespread{1.2}
 
@@
 
-\title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}
-\author[1]{Elisabeth Millington}
-\author[2]{Scott Powers}
-\affil[1]{Department of Kinesiology, Rice University}
-\affil[2]{Department of Sport Management, Rice University}
-
 \begin{document}
 
-\maketitle
+\title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}
+
+\author{Elisabeth Millington\affilnum{1} and Scott Powers\affilnum{2}}
+
+\affiliation{\affilnum{1}Department of Kinesiology, Rice University\\
+\affilnum{2}Department of Sport Management, Rice University}
+
+\corrauth{INSERT HERE}
+
+\email{INSERT HERE}
 
@@
 
+\keywords{INSERT 4-5 KEYWORDS HERE}
+
+\maketitle
+
 \section{Introduction}
-
 Random forests are a powerful machine learning prediction model that use the ensemble learning technique of combining multiple decision trees to mitigate overfitting.
@@
 \subsection{Related Work}
-
 Previous research has been done to attempt to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
@@
 \label{sec:data}
-
 All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}).
@@
 \label{sec:methods}
-
 \subsection{Random Forest}
-
 Random forests are a machine learning method typically used for classification or regression (\cite{breiman_random_2001}).
@@
 \subsection{Similarity Score}
-
 \textcite{lin_random_2006} previously established the relationship between $k$-NN and the random forest algorithm.
@@
 \label{sec:results}
-
 \subsection{Random Forest}
-
 Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
@@
 \subsection{Case Study: 2025 NFL Draft}
-
 Table \ref{tab:top-ten} shows the adjusted QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
@@
 \label{sec:software}
-
 The primary contribution of the present work is less about the specific application to quarterback prospects in football and more about introducing the concept of extracting similarity scores from random forests to interpret the predictions.
@@
 \label{sec:discussion}
-
 We have demonstrated a methodology for producing similarity scores to help interpret the predictions of random forests.

### 2025-07-28 scripts/generate_results.R
re-oder prospects in side-by-side similarity table

@@ -441,15 +441,15 @@ ss_top_10_wide <- ss_top_10 %>%
   select(`Player.Name`, Similarity) %>%
   setNames(c("SS_Player", "SS_Similarity"))
 
-cbind(cw_top_10_wide, ss_top_10_wide, jd_top_10_wide, dg_top_10_wide) %>%
+cbind(cw_top_10_wide, dg_top_10_wide, ss_top_10_wide, jd_top_10_wide) %>%
   sputil::write_latex_table(
     file = "tables/side_by_side_similarity.tex",
     colnames = rep(c("Comp", "Score"), times = 4),
     prefix_rows = "
       \\multicolumn{2}{c|}{Cam Ward} &
+      \\multicolumn{2}{c|}{Dillon Gabriel} &
       \\multicolumn{2}{c|}{Shedeur Sanders} &
-      \\multicolumn{2}{c|}{Jaxson Dart} &
-      \\multicolumn{2}{c}{Dillon Gabriel}
+      \\multicolumn{2}{c}{Jaxson Dart}
     ",
     align = "lr|lr|lr|lr"
   )

### 2025-07-28 articles/arxiv/paper.tex
final edits on paper

@@
 We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors.
-A player's similarity scores can be viewed as quantifiable player comps, and the random forest's prediction can be recovered by taking a weighted average of the outcome variable among those comps, with weights determined by the similarity scores.
-We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating (QBR) per season.
-We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.
+A player's similarity scores can be viewed as quantifiable player comps, and we show that the random forest's prediction can be recovered as the weighted average of the outcome variable among those comps, using the similarity scores as weights.
+We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating, averaged across seasons.
+We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
 \end{abstract}
@@
 Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position (\cite{hughes_positional_2015}).
-Due to the nature of the position, trading for or signing a solid quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
+Due to the nature of the position, trading for or signing a top quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
 Since quarterbacks are highly sought after, talented college quarterbacks with perceived NFL starting potential are typically taken in the first or second round of the draft.
@@
 Random forests can be thought of as an adaptive $k$-NN algorithm (\cite{lin_random_2006}), with the nearness of neighbors adaptively determined to minimize prediction error.
-Our work is moivated by this connection between random forests and the traditional lens of player comps.
-In this paper, we develop methods and software for extracting similarity scores from fitted random forest models in such a way that the prediction for a new data point is the weighted average of the outcomes in the training set, weighted according to these similarity scores.
+Our work is motivated by this connection between random forests and the traditional lens of player comps.
+In this paper, we develop methods and software for extracting similarity scores from fitted random forest models such that the prediction for a new data point is the weighted average of the outcomes in the training set, weighted by these similarity scores.
 
@@
 In Section \ref{sec:results}, we interpret the predictions from the NFL quarterback draft prospect random forest model by ranking players from the training set according to their similarity scores induced by the adaptive nearest neighbors interpretation of random forests.
-In Section \ref{sec:software}, we describe the new R package rfcomp for calculating these similarity scores from fitted random forest objects.
+In Section \ref{sec:software}, we describe the new R package treecomp for calculating these similarity scores from fitted random forest objects.
 In Section \ref{sec:discussion}, we summarize the key findings.
@@
 
-Previous research has been done to attempt to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
+Previous research has attempted to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
 \textcite{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
@@
 
-Draft models are fairly common in professional sports, as evaluating prospect talent can be key to success.
+Draft models are common in professional sports, as evaluating prospect talent can be key to success.
 One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \textcite{craig_predicting_2021}.
@@
 \textcite{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
-They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting a quarterback.
+They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks.
 Again, the logistic regression model results in predictions that are easily interpretable.
@@
 While past work in sport analytics has attempted to predict draft prospect outcomes using interpretable methods (i.e.
-logistic regression) or machine learning methods with greater predictive accuracy, the novelty of the current work is that we achieve interpretability with a machine learning moethod.
+logistic regression) or machine learning methods with greater predictive accuracy, the novelty of the current work is that we achieve interpretability with a machine learning method.
 In particular, we tie our novel interpretation of random forest predictions to the concept of comps, which is well understood by scouts and executives in the sports industry.
-The implementation of our method is available in the treecomp R package, allowing practitioners to easily apply it to their own fitted random forests.
+The implementation of our method is available in the R package treecomp, allowing practitioners and researchers to easily apply it to their own fitted random forests.
 
@@
 All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}).
-Our dataset includes 2,099 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season.
+Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their careers; and (3) attempted more than five passes per season.
 The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
@@
 The range of QBR is from 0 to 100, and \textcite{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
-For each quarterback, we calculate their average QBR per season, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero.
+For each quarterback, we calculate their average QBR across seasons, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero.
 Here the scale of QBR is helpful because players who reach the NFL cannot be rated below players who never reach the NFL.
@@
 Because of the date ranges of the data, we do not calculate QBR perfectly for all of the quarterbacks in our dataset.
-On one extremes, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
+On one extreme, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
 These quarterbacks have a QBR of zero in our dataset.
@@
 
-Another limitation of average QBR per season is that some players put up very high QBR numbers in very small samples (e.g.
-a QBR of 64.7 on only 26 passing attempts over a full season) and are rated as having very positive outcomes although a more subjective evaluation of their careers would be less positive.
+Another limitation of average QBR per season is that some players put up very high QBR numbers in very small samples (e.g., a QBR of 64.7 on only 26 passing attempts over a full season) and are rated as having very positive outcomes although a more subjective evaluation of their careers would be less positive.
 To mitigate the effect of this phenomenon, we regress each player's career QBR per season toward zero:
@@
 \end{equation*}
-We chose $c = 20.8$ as this value minimizes the error in predicting a player's next-season QBR when used to regrees current-season QBR.
+We chose $c = 20.8$ as this value minimizes the error in predicting a player's next-season QBR when used to regress current-season QBR.
 
@@
 Random forests are a machine learning method typically used for classification or regression (\cite{breiman_random_2001}).
-The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision within each bootstrap sample, randomly choosing at each node a subset of the features to consider for the split.
-The random forest's prediction is the average of the predictions of the individual treees.
+The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision tree on each bootstrap sample, randomly choosing at each split a subset of features as candidates.
+The random forest's prediction is the average of the predictions of the individual trees.
 By fitting multiple decision trees, random forests generally achieve much lower-variance predictions than individual decision trees.
@@
 
-We trained the random forest model using the R package ranger (\cite{wright_ranger_2015}), which is a fast implementation of the random forest algorithm.
-The response variable was average QBR per season (zero for players who never appeared in the NFL).
-The college statistics used as predictor variables were games played per season, completed passes per season, attempted passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
+We trained the random forest model using the R package ranger (\cite{wright_ranger_2024}), which is a fast implementation of the random forest algorithm.
+The response variable was average QBR across NFL seasons (zero for players who never appeared in the NFL).
+The college statistics used as predictor variables were games played per season, attemtped passes per season, complete passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
 
@@
 We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement.
-To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2018}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
-After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 5 and the optimal minimum node size to be 73.
+To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2024}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
+After 100 iterations of the tuning algorithm, we estimated the optimal number of candidate variables at each split to be 5 and the optimal minimum node size to be 73.
 
@@
 
-\textcite{lin_random_2006} previously established the relationship between $k$-NN and the random forest algorithm.
+Our method is motivated by the observation that a random forest is effectively $k$-NN, using neighborhoods adaptively determined to optimize predictive accuracy (\cite{lin_random_2006}).
 In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data.
@@
 We use $|\mathcal{T}_{b,i}|$ to denote the number of training observations (with repetition) which belong to $\mathcal{T}_{b,i}$.
-If observation $i$ is not included in bootstrap sample $b$, then $n_{b,i} = 0$, and we define $\mathcal{T}_{b,i} = \{i\}$ and $|\mathcal{T}_{b,i}| = 1$.
+If observation $i$ is not included in bootstrap sample $b$, then $n_{b,i} = 0$, and we define $\mathcal{T}_{b,i}$ as the terminal node to which observation $i$ would be assigned, according to the rules of decision tree $b$.
 
@@
 $$
-where $w_{b,0,i} = n_{b,i} \cdot \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ which comes from observation $i$.
-The weights $w_{b,\cdot,0}$ can be thought of as defining a local neighborhood around $\vec x_0$ for tree $b$.
+where $w_{b,0,i} = \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} \cdot n_{b,i} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ from observation $i$.
+The weights $w_{b,0,\cdot}$ can be interpreted as defining a local neighborhood around $\vec x_0$ for tree $b$ (\cite{lin_random_2006}).
 
@@
 
+The similarity score $\bar w_{0,i}$ can be thought of as describing how similar is observation 0 to observation $i$, with similarity determined adaptively in a meaningful way so as to optimize random forest predictions.
+In contrast, defining similarity based on Euclidean distance would not put more weight on features which are more important for predicting outcomes.
+As demonstrated by (\ref{eqn:rf-as-knn}), the similarity score $\bar w_{0,i}$ comes with the nice property that the prediction for $\vec x_0$ is recovered by calculating the weighted average of the outcome values in the training data, using $\bar w_{0,i}$ as the weights.
 Note that the similarity score is not symmetric.
 In general, $\bar w_{0,i} \ne \bar w_{i,0}$.
-In fact, as we have defined it, $\bar w_{i,0}$ is undefined if $\vec x_0$ is not in the training set.
-The similarity score $\bar w_{0,i}$ can be thought of as describing how similar is observation 0 to observation $i$, with similarity determined adaptively in a meaningful way so as to optimize random forest predictions.
-In contrast, defining similarity based on Euclidean distance would not put more weight on features which are more important for predicting outcomes.
-The similarity score $\bar w_{0,i}$ comes with the nice property that the prediction for $\vec x_0$ is recovered by calculating the weighted average of the outcome values in the training data, using $\bar w_{0,i}$ as the weights.
 
@@
 Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model.
-The model's test RMSE was 8.61 units of QBR, corresponding to explaining 43.7\% of the variance of QBR in the test set.
+The model's test RMSE was 8.61 units of QBR, corresponding to explaining 43.7\% of the variance of regressed QBR in the test set.
 Final-season Heisman voting, average passing performance (yards, touchdowns, completions) per season, and final-season strength of schedule stand out as the most important predictors of NFL success.
@@
 
-\begin{figure}[H]
+\begin{figure}
     \centering
-    \includegraphics[width=0.75\linewidth]{../../figures/variable_importance.pdf}
-    \caption{\textit{Variable importance plot for random forest trained to predict adjusted NFL QBR for college quarterbacks.}}
+    \includegraphics[width=0.5\linewidth]{../../figures/variable_importance.pdf}
+    \caption{\it Importance of predictor variables in random forest model. The response variable in this model is regressed NFL QBR, averaged across seasons in a player's career.}
     \label{fig:importance}
@@
 
-Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted adjusted QBR against career passing yards per season and final-season passer rating.
-These are two of the most important predictors in the model, and they have a relatively low correlation.
+Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest, plotting predicted QBR against career passing yards per season and final-season strength of schedule.
+Stength of schedule is a proxy for the level of competition at which the quarterback played, and we observe that the top predictions correspond to quarterbacks who had great success at high levels of competition.
 The extreme outlier by passing yards per season is Bailey Zappe, who played only one FBS season at Western Kentucky in Conference USA and was selected by the New England Patriots in the fourth round of the 2022 NFL Draft.
-The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed an adjusted QBR of 40.
+The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed a QBR of 40.
 This speaks to the large number of college quarterbacks who achieve zero QBR in the NFL and the difficulty of distinguishing between prospects based on college performance statistics alone.
@@
     \centering
-    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
-    \caption{\textit{Predicted adjusted QBR plotted against average passing yards per season and final-season passer rating.}}
+    \includegraphics[width=0.5\linewidth]{../../figures/3d_plot.pdf}
+    \caption{\it Visualization of prediction function learned by random forest model.
+Passing yards per season and final-season strength of schedule are two of most important predictor variables in the model.
+Each point represents one player from the historical training set, and the color corresponds to the player's fitted value.}
     \label{fig:3d-plot}
@@
 Figure \ref{fig:predicted-vs-actuals} shows actual vs.
-predicted adjusted QBR for all quarterbacks in the dataset.
+predicted QBR for all quarterbacks in the dataset.
 Recall that 91\% of actual QBRs are zero, which the visualization does not make clear.
-The errors are very right-skewed, and this illustrates why no college quarterback exceed an adjusted QBR prediction of 50.
+The errors are very right-skewed, and this further shows why so few college quarterbacks exceed a QBR prediction of 40.
 
@@
   \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
-  \caption{\textit{Predicted (from random forest) and actual adjusted QBR for all quarterbacks in the dataset.
-The vats majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}}
+  \caption{\it Predicted and actual (regressed) QBR for all quarterbacks in the dataset.
+The vast majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}
   \label{fig:predicted-vs-actuals}
@@
 
-Table \ref{tab:top-ten} shows the adjusted QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
-Cam Ward is predicted to have the highest adjusted QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft.
-Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest adjusted QBRs, respectively.
-While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025.
+Table \ref{tab:top-ten} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft.
+Cam Ward is predicted to have the highest QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft.
+Dillon Gabriel and Shedeur Sanders were projected to have the second- and third-highest QBRs, respectively.
+While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest of any quarterback in Heisman voting in 2024.
+Interestingly, Gabriel (third round) and Sanders (fifth round) were both selected by the Cleveland Browns after Sanders slid surprisingly deep into the draft, as we discuss later.
 Jaxson Dart's draft stock increased substantially after the Scouting Combine and his pro day.
@@
   \input{../../tables/top_ten.tex}
-  \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class.
-This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
+  \caption{\it Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class.
+This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}
   \label{tab:top-ten}
@@
 
-The predictions are interesting, but the focus of the present work is interpreting the predictions.
-Table \ref{tab:side-by-side-similarity} reports the ten most similar historical draft prospects for the top four quaterback prospects: Cam Ward, Shedeur Sanders, Jaxson Dart and Dillon Gabriel.
+While the predictions are interesting by themselves, the focus of the present work is interpreting how we arrive at them.
+Table \ref{tab:side-by-side-similarity} reports the ten most similar historical draft prospects for the top four quaterback prospects: Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
 Recall that the similarity scores reported in this table are interpretable as the weight of each historical prospect in the predictions in Table \ref{tab:top-ten}.
-Football fans will notice many household names among the comps for Ward, Sanders and Gabriel, but the comps for Dart are less impressive.
-Of note, the highest similarity score for any of these prospects is only 3.0\%.
-This is a helpful reminder that quarterback career outcomes are noisy, and the random forest model learns to spread out the weight across historical players, rather than concentrating a lot of weight in the single most similar player.
+The highest similarity score for any of these prospects is only 2.3\%.
+This is a helpful reminder that quarterback career outcomes are noisy and speaks to the fallacy of putting significant stock in a single player comparison.
+The random forest model learns to spread out the weight across historical players, rather than concentrating a lot of weight in the single most similar player.
+With access to more granular athleticism data, the weight might concentrate more in fewer comps.
 
@@
   }
-  \caption{\textit{Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft.
-The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}}
+  \caption{\it Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft.
+The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}
   \label{tab:side-by-side-similarity}
@@
 As a result, their college statistics used in the model do not include their statistics from these schools.
-We observe that Ward has the highest adjusted QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
+We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class.
 Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect.
-However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft.
+However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado.
 Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration.
-He also performed well in his interviews with teams, which is not a quantifiable measure.
+He also performed well in his interviews with teams (\cite{traina_qb_2025}), which is not quantifiable.
 
-Figure \ref{fig:prospect-plots} provides a full breakdown of the random forest predictions based on the each prospect's similarity scores for all of the historical prospects in the training data.
-Each prospect's prediction is a weighted average of past prospect's outcomes, and the weighted histograms in Figure \ref{fig:prospect-plots} visualize that weighted mean.
-We observe that all four 2025 prospects have a large concentration of weight on training players with 0 QBR, a reminder that even top prospects have a danger of becoming busts.
+Figure \ref{fig:prospect-plots} provides a full breakdown of the random forest predictions based on each prospect's similarity scores for all of the historical prospects in the training data.
+Each prospect's prediction is a weighted average of past prospects' outcomes, and the weighted histograms in Figure \ref{fig:prospect-plots} visualize that weighted mean.
+We observe that all four of the top quarterback prospects in 2025 have a large concentration of weight on training players with zero QBR, a reminder that even top prospects have a danger of becoming busts.
 Ward has the highest concentration of comps between 55 and 65 QBR, which is why his prediction is the highest.
-On the opposite end of the spectrum, Dart has the highest concentration of training players near 0 QBR, explaining why his prediction was by far the lowest of the four prospects.
+On the opposite end of the spectrum, Dart has the highest concentration of training players near zero QBR, explaining why his prediction was by far the lowest of the four prospects.
 
-\begin{figure}[H]
+\begin{figure}[h]
     \centering
@@
     \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
-    \caption{\textit{Top: Histogram of training players' adjusted QBR values weighted by similarity to the reference prospect.
-The vertical green line annotates each prospect's predicted adjusted QBR, which matches the mean of the histograph distribution.
-Bottom: Scatter plot showing individual similarity scores and adjusted QBR for historical prospects.}}
+    \caption{\it Top: Histogram of training players' regressed QBR values weighted by similarity to the prospect being predicted.
+The vertical green line annotates each prospect's predicted QBR, which matches the mean of the weighted distribution.
+Bottom: Scatter plot showing individual similarity scores and regressed QBR for historical prospects.}
     \label{fig:prospect-plots}
@@
 If the reference data are the same as the training data, then the predictions of the random forest can be recovered as the weighted average of the training outcomes, weighted by similarity score.
-However, it is also possible to use a different dataset as the reference data; in this case, the similarity is still interpretable as corresponding to the adaptive nearness learned by the random forest.
-This adaptive nearness is more informative than Euclidean distance because it reflects nearness in ways that are important for predicting outcomes.
+However, it is also possible to use a different dataset as the reference data; in this case, the similarity is still interpretable as reflecting the adaptive nearness learned by the random forest, which is more informative than a similarity based on Euclidean distance giving equal weight to all features.
 
-The treecomp R package is available at \url{https://github.com/elisabethmill/treecomp}.
+The treecomp package is available at \url{https://github.com/elisabethmill/treecomp}.
 
@@
 These similarity scores have the nice property that the random forest predictions are the weighted averages of outcomes in the training set, weighted according to these similarity scores.
-In addition to interpreting the random forest predictions, the methodology is helpful for identifying similar players based on the random forest as an adaptive $k$-NN algorithm, meaning that the neighborhoods are chosen in a meaningful way for predicting outcomes, which is not the case for a standard $k$-NN algorithm based on Euclidean distance.
+In addition to interpreting the random forest predictions, the methodology is helpful for identifying similar players based on the interpretation of a random forest as an adaptive $k$-NN algorithm.
+The neighborhoods are chosen in a meaningful way for predicting outcomes, which is not the case for a standard $k$-NN algorithm based on Euclidean distance.
 Our methodology is not limited to quarterbacks, to football, or even to sports, although the interpretation of predictions as a weighted average of historical comps fits well with the tradition of using comps in traditional scouting for sports.
 
-In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in adjusted QBR, which highlights some limitations of statistical forecasting when scouting draft prospects.
-The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance.
-However, these two things also highlight the importance of non-quantifiable factors---such as attitude, coachability, off-the-field issues, adaptability, and leadership.
+In the application to the 2025 NFL draft, the random forest's predictions explain slightly less than half of the out-of-sample variance in regressed QBR, which highlights some limitations of statistical forecasting when scouting draft prospects.
+The variable importance scores show which college statistics are most predictive of better NFL performance.
 
-[INSERT PARAGRAPH ABOUT SOFTWARE]
-
-[INSERT PARAGRAPH ABOUT FUTUR WORK EXTENDING TO OTHER TREE-BASED METHODS]
+The application of predicting the future success of NFL quarterback prospects is an illustration of our methodology, but the primary purpose of this work is to introduce the idea of extracting similarity scores from fitted random forest objects to explain the predictions using the framework of player comps, well known in sports.
+Importantly, we have published a user-friendly R package called treecomp, which will allow researchers and practitioners to extract similarity scores from their own random forest models.
+The R package name and the title of this paper refer more generally to tree-based methods, rather than specifically to random forests, and this is intentional.
+Future work could extend our methodology to other tree-based regression and classification models, such as gradient boosting.
 
@@
 
+\section*{Acknolwedgments}
+
+The authors thank Kevin Meers for suggestions that led to improvements in the random forest model for predicting NFL quarterback prospect success.
+
+\section*{Code and Data Availability}
+
+The data used for this paper are included in the R package treecomp, which is available at \url{https://github.com/elisabethmill/treecomp}.
+The repository also includes code for reproducing the results in this paper.
+
 \printbibliography

### 2025-07-28 scripts/generate_results.R
Polish paper before arXiv submission (#6)
* store SOS as numeric

* choose more interesting variables for #D plot

* fix latex results table formatting

* update latest results in paper

* fix mckenna reference with data

* re-oder prospects in side-by-side similarity table

* polish references

* final edits on paper

@@ -165,20 +165,19 @@ new_plot_data <- past_data %>%
 {
   sputil::open_device("figures/3d_plot.pdf", height = 5)
   plot <- new_plot_data |>
-    ggplot(aes(x = ncaa_yds_per_year, y = ncaa_passer_rating_last, color = predictions)) +
+    ggplot(aes(x = ncaa_yds_per_year, y = ncaa_sos_last, color = predictions)) +
     geom_point(size = 1) +
     scale_color_viridis_c() +
     labs(
       x = "Passing Yards per Season (College)",
-      y = "Final-Season Passer Rating (College)",
+      y = "Strength of Schedule (College)",
       color = "Predicted QBR"
     ) +
-    coord_cartesian(xlim = c(0, 6000), ylim = c(0, 250)) +
     theme_minimal() +
-    theme(legend.position = "inside", legend.position.inside = c(0.85, 0.25))
+    theme(legend.position = "inside", legend.position.inside = c(0.85, 0.75))
   print(plot)
   dev.off()
-  }
+}
 
 {
   sputil::open_device("figures/predicted_vs_actuals.pdf", height = 5, width = 5)
@@ -210,7 +209,11 @@ present_data %>%
   arrange(-predictions) %>%
   head(10) %>%
   mutate(predictions = sprintf("%.1f", predictions)) %>%
-  sputil::write_latex_table(file = "tables/top_ten.tex")
+  sputil::write_latex_table(
+    file = "tables/top_ten.tex",
+    colnames = c("Player", "Predicted QBR"),
+    align = "l|r"
+  )
 
 
 # Produce similarity figures and tables ----
@@ -438,5 +441,15 @@ ss_top_10_wide <- ss_top_10 %>%
   select(`Player.Name`, Similarity) %>%
   setNames(c("SS_Player", "SS_Similarity"))
 
-cbind(cw_top_10_wide, ss_top_10_wide, jd_top_10_wide, dg_top_10_wide) %>%
-  sputil::write_latex_table(file = "tables/side_by_side_similarity.tex")
+cbind(cw_top_10_wide, dg_top_10_wide, ss_top_10_wide, jd_top_10_wide) %>%
+  sputil::write_latex_table(
+    file = "tables/side_by_side_similarity.tex",
+    colnames = rep(c("Comp", "Score"), times = 4),
+    prefix_rows = "
[…]

### 2025-07-29 articles/arxiv/paper.tex
move figures and tables to arxiv folder (#7)

@@
 \affil[2]{Department of Sport Management, Rice University}
+\date{August 1, 2025}
 
@@
     \centering
-    \includegraphics[width=0.5\linewidth]{../../figures/variable_importance.pdf}
+    \includegraphics[width=0.5\linewidth]{figures/variable_importance.pdf}
     \caption{\it Importance of predictor variables in random forest model. The response variable in this model is regressed NFL QBR, averaged across seasons in a player's career.}
@@
     \centering
-    \includegraphics[width=0.5\linewidth]{../../figures/3d_plot.pdf}
+    \includegraphics[width=0.5\linewidth]{figures/3d_plot.pdf}
     \caption{\it Visualization of prediction function learned by random forest model.
@@
   \centering
-  \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
+  \includegraphics[width=0.5\linewidth]{figures/predicted_vs_actuals.pdf}
   \caption{\it Predicted and actual (regressed) QBR for all quarterbacks in the dataset.
@@
   \centering
-  \input{../../tables/top_ten.tex}
+  \input{tables/top_ten.tex}
   \caption{\it Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class.
@@
   \resizebox{\textwidth}{!}{
-    \input{../../tables/side_by_side_similarity.tex}
+    \input{tables/side_by_side_similarity.tex}
   }
@@
     \centering
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_ward.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_sanders.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_gabriel.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_dart.pdf} \\
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_ward.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_sanders.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_gabriel.pdf}
-    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_dart.pdf} \\
+    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_dart.pdf}
     \caption{\it Top: Histogram of training players' regressed QBR values weighted by similarity to the prospect being predicted.

### 2025-07-29 articles/arxiv/tables/side_by_side_similarity.tex
move figures and tables to arxiv folder (#7)

\begin{tabular}{lr|lr|lr|lr}

      \multicolumn{2}{c|}{Cam Ward} &
      \multicolumn{2}{c|}{Dillon Gabriel} &
      \multicolumn{2}{c|}{Shedeur Sanders} &
      \multicolumn{2}{c}{Jaxson Dart}
    \\
Comp & Score & Comp & Score & Comp & Score & Comp & Score \\
% latex table generated in R 4.5.0 by xtable 1.8-4 package
% Mon Jul 28 15:03:22 2025
  \hline
Johnny Manziel & 2.3\% & Mason Rudolph & 2.0\% & Mason Rudolph & 1.9\% & Teddy Bridgewater & 1.9\% \\ 
  Marcus Mariota & 2.3\% & Dwayne Haskins & 1.9\% & Kenny Pickett & 1.9\% & Tajh Boyd & 1.6\% \\ 
  Baker Mayfield & 2.2\% & Kenny Pickett & 1.9\% & Philip Rivers & 1.9\% & Russell Wilson & 1.5\% \\ 
  Philip Rivers & 2.1\% & C.J. Stroud & 1.9\% & Ben Roethlisberger & 1.8\% & John Beck & 1.5\% \\ 
  Trevor Lawrence & 2.0\% & Andrew Luck & 1.9\% & Case Keenum & 1.8\% & Drake Maye & 1.4\% \\ 
  Matt Leinart & 2.0\% & Bo Nix & 1.9\% & Russell Wilson & 1.8\% & Zach Terrell & 1.4\% \\ 
  Russell Wilson & 2.0\% & Trevor Lawrence & 1.9\% & Trevone Boykin & 1.8\% & Kevin Hogan & 1.2\% \\ 
  Lamar Jackson & 1.9\% & Philip Rivers & 1.8\% & Kellen Moore & 1.7\% & Blake Bortles & 1.2\% \\ 
  Deshaun Watson & 1.9\% & Aaron Rodgers & 1.7\% & Andrew Luck & 1.7\% & Sam Howell & 1.2\% \\ 
  Case Keenum & 1.9\% & Bryce Young & 1.6\% & C.J. Stroud & 1.7\% & Patrick Mahomes & 1.1\% \\ 
\end{tabular}

### 2025-08-01 articles/jsa/JSASubmission.tex
upload JSA submission files

--- 2025-07-29 articles/arxiv/paper.tex
@@
-\documentclass{article}
+\documentclass[Review, sageh, times]{sagej}
 
-\usepackage[alldates=year, block=space, style=apa]{biblatex}
-\setlength\bibitemsep{0.5\baselineskip}
-\addbibresource{../random-forest-comps.bib}
-
-\usepackage[margin=1in]{geometry}
-\linespread{1.2}
+\usepackage{amsmath, amsfonts, array, float, graphicx, subcaption, moreverb, url, natbib}
 
@@
 
-\usepackage{amsmath, amsfonts, array, authblk, float, graphicx, subcaption}
+\begin{document}
+
+\runninghead{Millington and Powers}
 
 \title{A comps-based approach for interpreting tree-based predictions with an application to the NFL draft}
-\author[1]{Elisabeth Millington}
-\author[2]{Scott Powers}
-\affil[1]{Department of Kinesiology, Rice University}
-\affil[2]{Department of Sport Management, Rice University}
-\date{August 1, 2025}
 
-\begin{document}
+\author{Elisabeth Millington\affilnum{1} and Scott Powers\affilnum{2}}
 
-\maketitle
+\affiliation{\affilnum{1}Department of Kinesiology, Rice University\\
+\affilnum{2}Department of Sport Management, Rice University}
+
+\corrauth{Scott Powers, 6100 Main St, Houston, TX, 77005}
+
+\email{scott.powers@rice.edu}
 
@@
 We can define interpretable similarity scores that quantify how much each training observation contributes to a given prediction by leveraging the connection between random forests and adaptive nearest neighbors.
-A player's similarity scores can be viewed as quantifiable player comps, and we show that the random forest's prediction can be recovered as the weighted average of the outcome variable among those comps, using the similarity scores as weights.
-We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating, averaged across seasons.
-We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
+A player's similarity scores can be viewed as quantifiable player comps, and the random forest's prediction can be recovered by taking a weighted average of the outcome variable among those comps, with weights determined by the similarity scores.
+We apply this methodology to evaluate quarterback prospects in the 2025 National Football League Draft using ESPN's Total Quarterback Rating (QBR) per season.
+We show that our approach captures meaningful structure in the data while providing interpretability, allowing us to identify comps for top prospects such as Cam Ward, Jaxson Dart, Shedeur Sanders, and Dillon Gabriel.
 \end{abstract}
+
+\keywords{National Football League, interpretable machine learning, draft projections, player comps, random forests, sports analytics}
+
+\maketitle
 
@@
 Each tree is built from a random bootstrap of the data, and the final prediction is determined by either majority tree vote (classification tasks) or averaging the tree responses (regression tasks).
-Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data (\cite{breiman_random_2001}), which makes them suitable for many real-world applications.
+Random forests have the flexibility to model many different functional relationships while maintaining robustness to noise and outliers within the data \citep{breiman_random_2001}, which makes them suitable for many real-world applications.
 
 Like other so-called ``black box'' machine learning algorithms, one drawback of the random forest is that the reasons for an individual's prediction can be difficult to interpret.
-Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction (\cite{ribeiro_why_2016}).
+Interpretability is crucial when using a machine learning algorithm because it allows the user to choose the best prediction method and it builds trust for the user in the prediction \citep{ribeiro_why_2016}.
 This is especially important in sports contexts because executives need to consider non-quantifiable information in their decision-making.
@@
 Perhaps one of the most important decisions that sports executives make is selecting quarterbacks in the National Football League (NFL) draft.
-Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position (\cite{hughes_positional_2015}).
+Quarterbacks are the backbone of offensive schemes for every NFL team and are arguably the most important position \citep{hughes_positional_2015}.
 Due to the nature of the position, trading for or signing a top quarterback during free agency is uncommon, which is why drafting a quarterback with high NFL potential is crucial for team success.
@@
 In 2025, the slotted contract for No.
-1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus (\cite{badenhausen_nfl_2025}).
+1 pick Cam Ward was \$48.4 million over four years, of which \$32.2 million was guaranteed as signing bonus \citep{badenhausen_nfl_2025}.
 
 In traditional scouting for sports, draft prospect discussions and predictions are often based on comparing a current prospect with similar draft prospects in the past.
-Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success (\cite{trapasso_nfl_2025}).
+Some of these past draft prospects will have gone on to successful careers, and others will have experienced less success \citep{trapasso_nfl_2025}.
 Scouts, managers, and coaches can assess NFL potential by comparing prospects to past players with similar collegiate careers.
-These historical reference players are known colloquially as ``comps'' (\cite{jones_nfl_2025}), short for comparables.
+These historical reference players are known colloquially as ``comps'' \citep{jones_nfl_2025}, short for comparables.
 The simplest example of this line of reasoning is, ``This prospect reminds me most of Player X, and Player X had a successful career, so I think this player will have a successful career,'' effectively a human implementation of the $k$-nearest neighbor ($k$-NN) algorithm, with $k = 1$.
 
-Random forests can be thought of as an adaptive $k$-NN algorithm (\cite{lin_random_2006}), with the nearness of neighbors adaptively determined to minimize prediction error.
+Random forests can be thought of as an adaptive $k$-NN algorithm \citep{lin_random_2006}, with the nearness of neighbors adaptively determined to minimize prediction error.
 Our work is motivated by this connection between random forests and the traditional lens of player comps.
@@
 Previous research has attempted to introduce interpretability to predictions from random forests and other tree-based machine learning algorithms.
-\textcite{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
-\textcite{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models.
-In a sports context, \textcite{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.
+\citet{petkovic_improving_2018} describe a user-centered approach to reporting random forest results in an interpretable manner.
+\citet{aria_comparison_2021} compare two different techniques for interpreting the interactions between predictor variables and response variables in random forest models.
+In a sports context, \citet{ouyang_integration_2024} use Shapley Additive exPlanations (SHAP) values to interpret NBA game outcome predictions from an XGBoost model.
 
 Draft models are common in professional sports, as evaluating prospect talent can be key to success.
-One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \textcite{craig_predicting_2021}.
+One example of a draft model that uses ESPN's Total Quarterback Rating as the response variable is \citet{craig_predicting_2021}.
 They used logistic regression to estimate whether a quarterback would be selected in the draft and used a second regression to estimate that quarterback's NFL performance based on their final season metrics, scout grades, and player height.
 Their predictions are easy to interpret directly from the logistic regression model.
-\textcite{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
+\citet{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
 They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks.
 Again, the logistic regression model results in predictions that are easily interpretable.
-\textcite{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
+\citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
 The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model.
-\textcite{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics.
+\citet{luo_improving_2024} propose a method for improving the way that NHL teams draft players by using both scouting reports and statistics.
 They use a combination of large language models and ensemble machine learning, which are both opaque in the way that they generate predictions, leading to uninterpretable predictions/reports.
@@
 
-All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}).
+All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}.
 Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their careers; and (3) attempted more than five passes per season.
@@
 
-In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) (\cite{burke_how_2016}) from each season between 2006 and 2024 (inclusive).
+In addition to college statistics, we obtained each player's passing attempts and Total Quarterback Rating (QBR) \citep{burke_how_2016} from each season between 2006 and 2024 (inclusive).
 Only available since 2006, QBR is based on expected points added (EPA), which is the change in drive scoring expectation before and after each play.
 It considers each quarterback's share of their team's EPA and accounts for home-field advantage, defensive strength, and garbage time.
-The range of QBR is from 0 to 100, and \textcite{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
+The range of QBR is from 0 to 100, and \citet{burke_how_2016} suggests thinking of it as a percentile of performance at the individual game level.
 For each quarterback, we calculate their average QBR across seasons, and we define a player who has attempted fewer than 10 career NFL passes to have a QBR of zero.
@@
 
-A response variable such as nflWAR (\cite{yurko_nflwar_2019}) or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers.
+A response variable such as nflWAR \citep{yurko_nflwar_2019} or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers.
 The primary purpose of the present work (extracting similarity scores to interpret random forest predictions through the lens of player comps) holds regardless of the response variable used.
@@
 
-Random forests are a machine learning method typically used for classification or regression (\cite{breiman_random_2001}).
+Random forests are a machine learning method typically used for classification or regression \citep{breiman_random_2001}.
 The algorithm creates $B$ bootstrap samples of the original data and then estimates a decision tree on each bootstrap sample, randomly choosing at each split a subset of features as candidates.
@@
 
-We trained the random forest model using the R package ranger (\cite{wright_ranger_2024}), which is a fast implementation of the random forest algorithm.
+We trained the random forest model using the R package ranger \citep{wright_ranger_2024}, which is a fast implementation of the random forest algorithm.
 The response variable was average QBR across NFL seasons (zero for players who never appeared in the NFL).
@@
 We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement.
-To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2024}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
+To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2024}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
 After 100 iterations of the tuning algorithm, we estimated the optimal number of candidate variables at each split to be 5 and the optimal minimum node size to be 73.
@@
 
-Our method is motivated by the observation that a random forest is effectively $k$-NN, using neighborhoods adaptively determined to optimize predictive accuracy (\cite{lin_random_2006}).
+Our method is motivated by the observation that a random forest is effectively $k$-NN, using neighborhoods adaptively determined to optimize predictive accuracy \citep{lin_random_2006}.
 In what follows, $\vec x_1, ..., \vec x_n \in \mathbb{R}^p$ represent the feature vectors in the training data, and $y_1, ..., y_n \in \mathbb{R}$ represent the outcome values in the training data.
@@
 where $w_{b,0,i} = \mathbb{I}\{\mathcal{T}_{b,0} = \mathcal{T}_{b,i}\} \cdot n_{b,i} / |\mathcal{T}_{b,0}|$ is the proportion of training data in $\mathcal{T}_{b,0}$ from observation $i$.
-The weights $w_{b,0,\cdot}$ can be interpreted as defining a local neighborhood around $\vec x_0$ for tree $b$ (\cite{lin_random_2006}).
+The weights $w_{b,0,\cdot}$ can be interpreted as defining a local neighborhood around $\vec x_0$ for tree $b$ \citep{lin_random_2006}.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.5\linewidth]{figures/variable_importance.pdf}
+    \includegraphics[width=0.7\linewidth]{variable_importance.pdf}
     \caption{\it Importance of predictor variables in random forest model. The response variable in this model is regressed NFL QBR, averaged across seasons in a player's career.}
@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.5\linewidth]{figures/3d_plot.pdf}
+    \includegraphics[width=0.7\linewidth]{3d_plot.pdf}
     \caption{\it Visualization of prediction function learned by random forest model.
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
-  \includegraphics[width=0.5\linewidth]{figures/predicted_vs_actuals.pdf}
+  \includegraphics[width=0.7\linewidth]{predicted_vs_actuals.pdf}
   \caption{\it Predicted and actual (regressed) QBR for all quarterbacks in the dataset.
@@
   \centering
-  \input{tables/top_ten.tex}
+  \input{top_ten.tex}
   \caption{\it Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class.
@@
 Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft.
-However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_2025}).
+However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance \citep{mckenna_what_2025}.
 In the end, Dart was the other quarterback taken in the first round of the draft.
@@
   \resizebox{\textwidth}{!}{
-    \input{tables/side_by_side_similarity.tex}
+    \input{side_by_side_similarity.tex}
   }
@@
 Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration.
-He also performed well in his interviews with teams (\cite{traina_qb_2025}), which is not quantifiable.
+He also performed well in his interviews with teams \citep{traina_qb_2025}, which is not quantifiable.
 
@@
     \centering
-    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_ward.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_sanders.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_gabriel.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_histogram_dart.pdf} \\
-    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_ward.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_sanders.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_gabriel.pdf}
-    \includegraphics[width=0.24\linewidth]{figures/prospect_similarity_dart.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_histogram_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_histogram_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_histogram_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_histogram_dart.pdf} \\
+    \includegraphics[width=0.24\linewidth]{prospect_similarity_ward.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_similarity_sanders.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_similarity_gabriel.pdf}
+    \includegraphics[width=0.24\linewidth]{prospect_similarity_dart.pdf}
     \caption{\it Top: Histogram of training players' regressed QBR values weighted by similarity to the prospect being predicted.
@@
 
-\section*{Acknolwedgments}
+\section{Acknolwedgments}
 
@@
 
-\section*{Code and Data Availability}
+\section{Author contributions}
+
+EM downloaded the data, calculated the similarity scores, and drafted the manuscript. SP provided the research idea, converted EM's code into an R package, and revised the manuscript.
+
+\section{Statements and Declarations}
+\subsection{Ethical Considerations}
+
+Not applicable
+
+\subsection{Consent to Participate}
+
+Not applicable
+
+\subsection{Consent for Publication}
+
+Not applicable
+
+\subsection{Declaration of Conflicting Interest}
+
+The authors declared no potential conflicts of interest with respect to the research, authorship, and/or publication of this article.
+
+\subsection{Funding Statement}
+
+The authors received no financial support for the research, authorship, and/or publication of this article.
+
+\subsection{Code and Data Availability}
 
@@
 
-\printbibliography
+\bibliographystyle{SageH}
+\bibliography{random-forest-comps}
 

### 2025-12-31 articles/jsa/JSASubmission.tex
revise JSA submission

@@
 This poses challenges for interpretability, particularly in environments where decisions can be very high-stakes, like professional sports.
-In this paper, we present a method for interpreting random forest predictions in a sports setting through the lens of player comps.
+In this paper, we \textcolor{red}{utilize the method of \cite{lin_random_2006}} for interpreting random forest predictions in a sports setting through the lens of player comps.
 Player comps are a common practice in player scouting, where prospects are evaluated based on their perceived similarity to past players.
@@
 The paper is organized as follows.
-In Section \ref{sec:data}, we describe a dataset on which we apply our methodology to evaluate NFL draft quarterback prospects, using ESPN's Total Quarterback Rating as the response variable and college statistics as the predictor variables.
-In Section \ref{sec:methods}, we derive similarity scores from a fitted random forest such that the model's predictions are equal to a similarity-weighted average of the training outcomes.
-In Section \ref{sec:results}, we interpret the predictions from the NFL quarterback draft prospect random forest model by ranking players from the training set according to their similarity scores induced by the adaptive nearest neighbors interpretation of random forests.
-In Section \ref{sec:software}, we describe the new R package treecomp for calculating these similarity scores from fitted random forest objects.
-In Section \ref{sec:discussion}, we summarize the key findings.
+In \textcolor{red}{Section \nameref{sec:data}}, we describe a dataset on which we apply our methodology to evaluate NFL draft quarterback prospects, using ESPN's Total Quarterback Rating as the response variable and college statistics as the predictor variables.
+In Section \nameref{sec:methods}, we derive similarity scores from a fitted random forest such that the model's predictions are equal to a similarity-weighted average of the training outcomes.
+In Section \nameref{sec:results}, we interpret the predictions from the NFL quarterback draft prospect random forest model by ranking players from the training set according to their similarity scores induced by the adaptive nearest neighbors interpretation of random forests.
+In Section \nameref{sec:software}, we describe the new R package treecomp for calculating these similarity scores from fitted random forest objects.
+In Section \nameref{sec:discussion}, we summarize the key findings.
 
@@
 \citet{wolfson_quarterback_2011} use their own metric of net points combined with games played to define NFL success.
-They estimate a logistic regression model for this metric on a combination of NFL Combine and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks.
+They estimate a logistic regression model for this metric on a combination of \textcolor{red}{NFL Draft Combine} and college statistics, and they use this model to analyze how effectively NFL teams leverage data available when drafting quarterbacks.
 Again, the logistic regression model results in predictions that are easily interpretable.
-\citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and combine data.
+\citet{berger_jumping_2021} use principal component analysis (PCA) combined with a high-order regression model to evaluate NBA prospects using their college statistics and \textcolor{red}{Draft Combine} data.
 The interpretation of the principal components is unclear, resulting in a relatively uninterpretable draft model.
@@
 All data used for this study are publicly available from Sports Reference \citep{sports_reference_sports_2025}.
-Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their careers; and (3) attempted more than five passes per season.
+Our dataset includes 2,099 quarterbacks who (1) played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive); (2) played more than six total games in their \textcolor{red}{college} careers; and (3) attempted more than \textcolor{red}{fifty passes per season}.
 The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).
@@
 In the dataset, 91\% of college quarterbacks achieve zero QBR in the NFL, so the distribution is very right-skewed.
-Because of the date ranges of the data, we do not calculate QBR perfectly for all of the quarterbacks in our dataset.
+\textcolor{red}{This percentage makes sense because let's say that each of the 136 FBS schools graduates one QB prospect every two years, resulting in 78 eligible prospects each year.
+9\% of these prospects translates to about 7 quarterbacks per year who will play in an NFL game at some point in their career (and not necessarily as a starter).
+Normally, the number of NFL quarterback vacancies hovers around 5 each year, leaving two players who will play at some point as backups.} Because of the date ranges of the data, we do not calculate QBR perfectly for all of the quarterbacks in our dataset.
 On one extreme, quarterbacks who finished their college careers in 2001 may have appeared in the NFL but only before 2006.
@@
 We chose $c = 20.8$ as this value minimizes the error in predicting a player's next-season QBR when used to regress current-season QBR.
+\textcolor{red}{The shrinkage parameter $c$ controls how strongly small-sample QBR values are pulled toward zero.
+We estimate $c$ using historical, year-to-year QBR data by choosing the value that minimizes the squared error between observed next-season QBR and the mean-regressed prediction:}
 
+\textcolor{red}{\begin{equation}
+  \hat{c}
+  =
+  \arg\min_{c}
+  \sum_i
+  \left(
+    \text{QBR}_{t+1,i}
+    -
+    \frac{\text{ATT}_{t,i}\,\text{QBR}_{t,i}}
+         {\text{ATT}_{t,i} + c}
+  \right)^2 .
+\end{equation}
+}
 A response variable such as nflWAR \citep{yurko_nflwar_2019} or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers.
@@
 The response variable was average QBR across NFL seasons (zero for players who never appeared in the NFL).
-The college statistics used as predictor variables were games played per season, attemtped passes per season, complete passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season Hesiman Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
+The college statistics used as predictor variables were games played per season, \textcolor{red}{attempted} passes per season, complete passes per season, passing yards per season, passing yards per attempt, passing touchdowns per season, passing interceptions per season, rushing attempts per season, rushing yards per season, rushing touchdowns per season, number of All-America selections, final-season \textcolor{red}{Heisman} Award voting finish, final-season strength of schedule, final-season games played, final-season passing yards per attempt, and final-season passer efficiency rating.
 
 To train the random forest model, we first implemented a random 70\%/30\% train/test split.
-We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement.
+We chose $B$ = 1,000 decision trees, each trained on a \textcolor{red}{bootstrap} sample of the data, with replacement.
 To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2024}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters.
@@
 
-The similarity score $\bar w_{0,i}$ can be thought of as describing how similar is observation 0 to observation $i$, with similarity determined adaptively in a meaningful way so as to optimize random forest predictions.
+\textcolor{red}{The score ranges from 0-1 and we express it later from 0\%-100\%.} The similarity score $\bar w_{0,i}$ can be thought of as describing how similar is observation 0 to observation $i$, with similarity determined adaptively in a meaningful way so as to optimize random forest predictions.
 In contrast, defining similarity based on Euclidean distance would not put more weight on features which are more important for predicting outcomes.
@@
 While the predictions are interesting by themselves, the focus of the present work is interpreting how we arrive at them.
-Table \ref{tab:side-by-side-similarity} reports the ten most similar historical draft prospects for the top four quaterback prospects: Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
+Table \ref{tab:side-by-side-similarity} reports the ten most similar historical draft prospects for the top four \textcolor{red}{quarterback} prospects: Cam Ward, Dillon Gabriel, Shedeur Sanders, and Jaxson Dart.
 Recall that the similarity scores reported in this table are interpretable as the weight of each historical prospect in the predictions in Table \ref{tab:top-ten}.
@@
 
-One key strength of this approach to comps is its objectivity.
+One key strength of this approach to comps is \textcolor{red}{that it is data-driven}.
 It enables teams to leverage historical data to make comps, without relying solely on scouting narratives or subjective analysis.
@@
 \bibliographystyle{SageH}
-\bibliography{random-forest-comps}
+\bibliography{articles/random-forest-comps}
 

### 2025-12-31 scripts/quantregForest_model.R
Upload the quantregForest model

@@ -1,0 +1,476 @@
+library(dplyr)
+library(ranger)
+library(tuneRanger)
+library(mlr)
+library(ggplot2)
+library(xtable)
+library(quantregForest)
+
+num_trees <- 1000
+
+set.seed(123)
+
+# Determine mean regression for QBR ----
+
+QBR2_passing_data <- treecomp::nfl_qbr_by_year %>%
+  arrange(player_name, year) %>%
+  group_by(player_name) %>%
+  mutate(
+    qbr_2 = lead(qbr),
+    qbr_2 = ifelse(is.na(qbr_2), 0, qbr_2)
+  ) %>%
+  filter(year != 2024)
+
+QBR2_model <- nls(
+  formula = qbr_2 ~ (att * qbr) / (att + c),
+  data = QBR2_passing_data,
+  start = list(c = 1)  # Starting guess for c
+)
+
+summary(QBR2_model)
+
+c <- coef(QBR2_model)[1]
+
+# Tune random forest ----
+
+data <- treecomp::quarterback %>%
+  mutate(
+    reg_qbr = nfl_att_per_year / (nfl_att_per_year + c) * nfl_qbr_per_year,
+    # NA for Heisman voting means you didn't finish in the top 10
+    ncaa_heisman_last = ifelse(is.na(ncaa_heisman_last), 11, ncaa_heisman_last)
+  )
+
+past_data <- data %>%
+  filter(
+    ncaa_year_last < 2024,
+    ncaa_games_career > 6,
+    ncaa_att_per_year > 50,
+    # Exclude players with very few NFL attempts per season because their QBR will be unstable
+    is.na(nfl_att_per_year) | (nfl_att_per_year == 0) | nfl_att_per_year >= 10
+  ) %>%
+  # Remaining NAs are due to not playing in the NFL
+  mutate_all(~ replace(., is.na(.), 0))
+
+set.seed(123)  # For reproducibility
+
+train_indices <- sample(1:nrow(past_data), size = 0.7 * nrow(past_data))
+train_data <- past_data[train_indices, ]
+test_data  <- past_data[-train_indices, ]
+
+predictor_vars <- c(
+  "ncaa_yds_per_att_career", "ncaa_games_per_year",
+  "ncaa_att_per_year", "ncaa_cmp_per_year", "ncaa_yds_per_year",
+  "ncaa_td_per_year", "ncaa_int_per_year",
[…]
